---
title: "07_preliminary-analyses"
author: "Anjum K. Gujral"
date: '2024-11-11'
output:
  html_document:
    df_print: paged
  pdf_document:
    output_dir: /Users/anjumgujral/Documents/repos/yuba-phys
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
trait_data1 <- read.csv("/Users/anjumgujral/Documents/repos/yuba-phys/trait_data1.csv")

install.packages("tinytex")
library('tinytex')
library('dplyr')
library('tidyverse')
library('ggplot2')
library(emmeans)
library(lme4)
library(lmerTest)
library(gridExtra)

```

#size distribution
```{r, echo = FALSE, message = FALSE, warning = FALSE}
# create a new column for size that categorizes individuals with diameter at root collar > 2.5 cm as saplings and individuals with diameter at root collar < 2.5 cm seedlings



ggplot(trait_data1, aes(x = DBH_complete, fill = size)) +
  geom_histogram(binwidth = 0.5, position = "dodge", alpha = 0.7, color = "black") +
  labs( 
       x = "Diameter at root collar (cm)", 
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "lightgreen")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

# microsite distributions
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = DBH_complete, y = percent_cover, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
       y = "Canopy closure", 
       x = "Diameter at root collar") +
  scale_color_manual(values = c("blue", "green", "red", "purple")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

Distribution of microsite environments across elevation
-- how similar are microsites at high and low elevation 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = percent_cover, fill = microsite)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Canopy Closure"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))

# Create a new variable combining size_class and elevation
trait_data1$size_elevation <- interaction(trait_data1$size, trait_data1$elevation)


ggplot(trait_data1, aes(x = species, y = percent_cover)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Canopy closure", 
    x = "Species",
    color = "Size & Elevation"  
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )


ggplot(trait_data1, aes(x = microsite, y = percent_cover)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Canopy closure", 
    x = "Microsite level", 
    color = "Size & Elevation"  
  ) +
  facet_wrap(~ species, scales = "free_y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```

# stand basal area
Distribution of stand basal area across elevation
-- how similar is stand density between plots at high and low elevation
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = stand_basal_area, fill = elevation)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Stand Basal Area"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))



ggplot(trait_data1, aes(x = species, y = stand_basal_area)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Stand Basal Area ", 
    x = "Species",
    color = "Size & Elevation"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )


ggplot(trait_data1, aes(x = microsite, y = stand_basal_area)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Stand Basal Area", 
    x = "Microsite level", 
    color = "Size & Elevation" 
  ) +
  facet_wrap(~ species, scales = "free_y") +
  
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```

# environmental variables
```{r}
# Create a correlation matrix for environmental variables
cor_matrix <- cor(trait_data1[, c("percent_cover", "elev_indiv", "stand_basal_area")], use = "pairwise.complete.obs")
print(cor_matrix)

cor(trait_data1$elev_indiv, trait_data1$stand_basal_area, 
    use = "complete.obs", method = "pearson")

cor.test(trait_data1$stand_basal_area, trait_data1$elev_indiv, method="pearson")

install.packages("ggcorrplot") 
library(ggcorrplot)
library(corrplot)
library("ggpubr")

# Create a custom set of axis labels (modify as per your needs)
custom_labels <- c("Canopy Closure", "Elevation (ft)", "Stand Basal Area")

colnames(cor_matrix) <- custom_labels
rownames(cor_matrix) <- custom_labels

# Visualize the correlation matrix with custom axis labels
corrplot(cor_matrix, 
         method = "color",          
         type = "lower",            
         order = "hclust",          
         addCoef.col = "black",    
         tl.col = "black",          
         tl.srt = 45,               
         tl.labels = custom_labels, 
         title = "Correlation Plot"
)



```

# Figure 1 - size
```{r}

s1 <- ggplot(trait_data1, aes(y = predawn_MPa_combined, x = diameter, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +  
  labs(
    x = "Diameter at root collar (cm)",
    y = "Predawn water potential (-MPa)",
    color = "Species"
  ) +
    theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )

s1

s2 <- ggplot(trait_data1, aes(x = diameter_complete, 
  y = -midday_MPa_combined, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  ylim (-3, -0.5)+
  labs(
    x = "Diameter at root collar (cm)", y = "Midday water potential (-MPa)", color = "Species") +
    theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )
s2

s3 <- ggplot(trait_data1, aes(x = diameter_complete, y = HSM_predawn_P50_mean, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) + 
  ylim (1,5) +
  labs(x = "Diameter at root collar (cm)", y = "HSM predawn (MPa)", color ="Species") +
    theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )
s3

s4 <- ggplot(trait_data1, aes(x = diameter_complete, y = HSM_midday_P50_mean, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  ylim (1,5) +
  labs(x = "Diameter at root collar (cm)", y = "HSM midday (MPa)", color ="Species") +
    theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )
s4

s5 <- ggplot(trait_data1, aes(x = diameter_complete, y = -P50_combined, color = species)) +
    geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  labs(x = "Diameter at root collar (cm)", y = "P50 (-MPa)", color ="Species") +
    theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )

s5
grid.arrange(s1, s2, s3, s4, s5, nrow = 3, ncol = 2)

```

## plotting model fit
```{r}
library(RColorBrewer)
library(ggnewscale)
library(colorspace)

species_colors <- setNames(
  brewer.pal(4, "Dark2"),
  c("ABCO", "ABMA", "PIPO", "PILA")
)


line_colors <- colorspace::darken(species_colors, amount = 0.1)


species_list <- list(
  "ABCO" = ABCO,
  "ABMA" = ABMA,
  "PIPO" = PIPO,
  "PILA" = PILA
)

species_list_elev <- list(
  "ABCO" = ABCO,
  "PIPO" = PIPO,
  "PILA" = PILA
)


### FIGURE 1A: Predawn by size
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(predawn_MPa_combined ~ diameter_complete_scaled + canopy_closure_scaled + (1|plot) + (1|year), data = df)
  } else {
    mod <- lmer(predawn_MPa_combined ~ diameter_complete_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot) + (1|year), data = df)
  }

  # Extract fixed-effect p-value for diameter
  summ <- summary(mod)$coefficients
  diam_p <- summ["diameter_complete_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  diam_p < 0.05 ~ "Significant",
  diam_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)


  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = diam_p,
                                                 signif = signif_label))

  # Compute mean and SD for scaling diameter
  mean_d <- mean(df$diameter_complete, na.rm = TRUE)
  sd_d   <- sd(df$diameter_complete, na.rm = TRUE)

  # Sequence of diameters
  diam_seq <- seq(min(df$diameter_complete, na.rm = TRUE),
                  max(df$diameter_complete, na.rm = TRUE),
                  length.out = 100)

  # Prediction data
  if (sp == "ABMA") {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_d)/sd_d,
      canopy_closure_scaled = 0
    )
  } else {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_d)/sd_d,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0
    )
  }

  newdata$diameter_cm <- diam_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### FIGURE 1A: Predawn WP by size
library(ggnewscale)

Fig1A <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = diameter_complete, y = predawn_MPa_combined, color = species),
             alpha = 0.35, size = 2.5, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  new_scale_color() +
  geom_line(
  data = all_preds,
  aes(x = diameter_cm, y = pred,
      color = species,
      linetype = signif),
  size = 1.2,
  na.rm = TRUE
)+
  scale_color_manual(values = species_colors) +
scale_linetype_manual(values = c(
  "Significant"     = "solid",
  "Marginal"        = "dashed",
  "Not significant" = "dotted"
)) +
scale_alpha_manual(values = c(
  "Significant"     = 1.0,
  "Marginal"        = 0.7,
  "Not significant" = 0.35
)) +
  scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
) +
  labs(
    x = "Basal diameter (cm)",
    y = expression(Psi[predawn]~"(MPa)")
  ) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  
    panel.grid = element_blank(),       
    axis.line = element_blank(),        
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )

Fig1A

### FIGURE 1B: Midday by size
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(midday_MPa_combined ~ diameter_complete_scaled + canopy_closure_scaled + (1|plot) + (1|year), data = df)
  } else {
    mod <- lmer(midday_MPa_combined ~ diameter_complete_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot) + (1|year), data = df)
  }

  # Extract fixed-effect p-value for diameter
  summ <- summary(mod)$coefficients
  diam_p <- summ["diameter_complete_scaled", "Pr(>|t|)"]
signif_label <- dplyr::case_when(
  diam_p < 0.05 ~ "Significant",
  diam_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)


  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = diam_p,
                                                 signif = signif_label))

  # Compute mean and SD for scaling diameter
  mean_d <- mean(df$diameter_complete, na.rm = TRUE)
  sd_d   <- sd(df$diameter_complete, na.rm = TRUE)

  # Sequence of diameters
  diam_seq <- seq(min(df$diameter_complete, na.rm = TRUE),
                  max(df$diameter_complete, na.rm = TRUE),
                  length.out = 100)

  # Prediction data
  if (sp == "ABMA") {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_d)/sd_d,
      canopy_closure_scaled = 0
    )
  } else {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_d)/sd_d,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0
    )
  }

  newdata$diameter_cm <- diam_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### FIGURE 1B: Midday WP by size
Fig1B <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = diameter_complete, y = midday_MPa_combined, color = species),
             alpha = 0.35, size = 2.5, na.rm = TRUE) +
 geom_line(
  data = all_preds,
  aes(x = diameter_cm, y = pred,
      color = species,
      linetype = signif),
  size = 1.2,
  na.rm = TRUE
) +
  scale_color_manual(values = species_colors) +
scale_linetype_manual(values = c(
  "Significant"     = "solid",
  "Marginal"        = "dashed",
  "Not significant" = "dotted"
)) + 
  scale_alpha_manual(values = c(
  "Significant"     = 1.0,
  "Marginal"        = 0.7,
  "Not significant" = 0.35
)) +  
  scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
) + 
  labs(x = "Basal diameter (cm)",
       y = expression(Psi[midday]~"(MPa)"))  +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  # ← this creates the box
    panel.grid = element_blank(),       # removes all grid lines
    axis.line = element_blank(),        # optional, can remove individual axis lines
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )

Fig1B
# Unified y-axis
ymin <- min(c(trait_data1$predawn_MPa_combined, trait_data1$midday_MPa_combined), na.rm = TRUE)
ymax <- max(c(trait_data1$predawn_MPa_combined, trait_data1$midday_MPa_combined), na.rm = TRUE)

Fig1A <- Fig1A + coord_cartesian(ylim = c(ymax, ymin))
Fig1B <- Fig1B + coord_cartesian(ylim = c(ymax, ymin))


grid.arrange(Fig1A, Fig1B, ncol = 2)

```

# Figure 2- P50

```{r}

r1 <- ggplot(trait_data1, aes(y = HSM_midday_P50_mean, x = predawn_MPa_combined, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +  
  labs(
    x = "Predawn Water Potential (-MPa)",
    y = "HSM Midday (MPa)",
    color = "Species"
  ) +
    theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )
r1


r2 <- ggplot(trait_data1, aes(y = -P50_combined, x = predawn_MPa_combined, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +  
  labs(
    x = "Predawn Water Potential (-MPa)",
    y = "P50 (-MPa)",
    color = "Species"
  ) +
    theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )

r2

m1 <- ggplot(trait_data1, aes(x = percent_cover, y = HSM_midday_P50_mean, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Canopy closure (%)", 
    y = "HSM predawn (MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse()  +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )

m1

m2 <- ggplot(trait_data1, aes(x = percent_cover, y = -P50_combined, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Canopy closure (%)", 
    y = "P50 (-MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )

m2

e1 <- ggplot(subset(trait_data1, species != "ABMA"), aes(x = elev_indiv, y = HSM_predawn_P50_mean, color = species))+
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "HSM predawn (MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )

e1

e2 <- ggplot(subset(trait_data1, species != "ABMA"), aes(x = elev_indiv, y = -P50_combined, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "P50 (-MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )

e2

grid.arrange(m2, m1, nrow = 1, ncol = 2)
grid.arrange(r1, r2,  m2, m1, e1, e2, nrow = 3, ncol = 2)

```
## plotting model fit

```{r}

### FIGURE 2A: P50 by size
all_preds_P50 <- data.frame()
model_summaries_P50 <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(P50_combined ~ diameter_complete_scaled + canopy_closure_scaled + (1|plot) + (1|year), data = df)
  } else {
    mod <- lmer(P50_combined ~ diameter_complete_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot) + (1|year), data = df)
  }

  # Extract fixed-effect p-value for diameter
  summ <- summary(mod)$coefficients
  diam_p <- summ["diameter_complete_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  diam_p < 0.05 ~ "Significant",
  diam_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)
  model_summaries_P50 <- dplyr::bind_rows(model_summaries_P50,
                                          data.frame(species = sp,
                                                     p_value = diam_p,
                                                     signif = signif_label))

  # Compute mean and SD for scaling diameter
  mean_d <- mean(df$diameter_complete, na.rm = TRUE)
  sd_d   <- sd(df$diameter_complete, na.rm = TRUE)

  # Sequence of diameters
  diam_seq <- seq(min(df$diameter_complete, na.rm = TRUE),
                  max(df$diameter_complete, na.rm = TRUE),
                  length.out = 100)

  # Prediction data
  if (sp == "ABMA") {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_d)/sd_d,
      canopy_closure_scaled = 0
    )
  } else {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_d)/sd_d,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0
    )
  }

  newdata$diameter_cm <- diam_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds_P50 <- dplyr::bind_rows(all_preds_P50, newdata)
}

### FIGURE 2A: P50
Fig2A <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = diameter_complete, y = P50_combined, color = species),
             alpha = 0.35, size = 2.5, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  new_scale_color() +
  geom_line(
  data = all_preds_P50,
  aes(x = diameter_cm, y = pred,
      color = species,
      linetype = signif),
  size = 1.2,
  na.rm = TRUE
)+
  scale_color_manual(values = species_colors) +
scale_linetype_manual(values = c(
  "Significant"     = "solid",
  "Marginal"        = "dashed",
  "Not significant" = "dotted"
)) +
scale_alpha_manual(values = c(
  "Significant"     = 1.0,
  "Marginal"        = 0.7,
  "Not significant" = 0.35
)) +
  scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
) +
  labs(x = "Basal diameter (cm)",
       y = expression(P[50]~"(MPa)"))+ 
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  
    panel.grid = element_blank(),       
    axis.line = element_blank(),         
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig2A

### FIGURE 2B: P50 by predawn 
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(P50_combined ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled +
                  (1|plot) + (1|year), data = df)
  } else {
    mod <- lmer(P50_combined ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled +
                  elev_indiv_scaled + (1|plot) + (1|year), data = df)
  }

  # Extract fixed-effect p-value for predawn
  summ <- summary(mod)$coefficients
  pd_p <- summ["predawn_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  pd_p < 0.05 ~ "Significant",
  pd_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)

  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = pd_p,
                                                 signif = signif_label))

  # Compute mean and SD for scaling predawn
  mean_pd <- mean(df$predawn_MPa_combined, na.rm = TRUE)
  sd_pd   <- sd(df$predawn_MPa_combined, na.rm = TRUE)

  # Sequence of predawn values
  pd_seq <- seq(min(df$predawn_MPa_combined, na.rm = TRUE),
                max(df$predawn_MPa_combined, na.rm = TRUE),
                length.out = 100)

  # Prediction data
  if (sp == "ABMA") {
    newdata <- data.frame(
      predawn_scaled = (pd_seq - mean_pd)/sd_pd,
      diameter_complete_scaled = 0,
      canopy_closure_scaled = 0
    )
  } else {
    newdata <- data.frame(
      predawn_scaled = (pd_seq - mean_pd)/sd_pd,
      diameter_complete_scaled = 0,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0
    )
  }

  newdata$predawn <- pd_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### FIGURE 2B: P50 by predawn
Fig2B <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = predawn_MPa_combined, y = P50_combined, color = species),
             alpha = 0.35, size = 2.5, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = predawn, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  new_scale_color() +
  scale_color_manual(values = species_colors) +
scale_linetype_manual(values = c(
  "Significant"     = "solid",
  "Marginal"        = "dashed",
  "Not significant" = "dotted"
)) +
scale_alpha_manual(values = c(
  "Significant"     = 1.0,
  "Marginal"        = 0.7,
  "Not significant" = 0.35
)) +
  scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
  ) +
  scale_x_reverse(
    labels = function(x) paste0("\u2212", abs(x))
  ) +
  labs(x = expression(Psi[predawn]~"(MPa)"),
       y = expression(P[50]~"(MPa)"))   +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  # ← this creates the box
    panel.grid = element_blank(),       
    axis.line = element_blank(),       
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )

Fig2B

### FIGURE 2C: P50_combined by percent cover
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(P50_combined ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled +
                  (1|plot) + (1|year), data = df)
  } else {
    mod <- lmer(P50_combined ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled +
                  elev_indiv_scaled + (1|plot) + (1|year), data = df)
  }

  # Extract fixed-effect p-value for canopy closure
  summ <- summary(mod)$coefficients
  cc_p <- summ["canopy_closure_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  cc_p < 0.05 ~ "Significant",
  cc_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)

  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = cc_p,
                                                 signif = signif_label))

  # Compute mean and SD for scaling percent cover
  mean_cc <- mean(df$percent_cover, na.rm = TRUE)
  sd_cc   <- sd(df$percent_cover, na.rm = TRUE)

  # Sequence of percent cover values
  cc_seq <- seq(min(df$percent_cover, na.rm = TRUE),
                max(df$percent_cover, na.rm = TRUE),
                length.out = 100)

  # Prediction data
  if (sp == "ABMA") {
    newdata <- data.frame(
      canopy_closure_scaled = (cc_seq - mean_cc)/sd_cc,
      diameter_complete_scaled = 0,
      predawn_scaled = 0
    )
  } else {
    newdata <- data.frame(
      canopy_closure_scaled = (cc_seq - mean_cc)/sd_cc,
      diameter_complete_scaled = 0,
      predawn_scaled = 0,
      elev_indiv_scaled = 0
    )
  }

  newdata$percent_cover <- cc_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### Plot
Fig2C <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = percent_cover, y = P50_combined, color = species),
             alpha = 0.35, size = 2.5, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = percent_cover, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  new_scale_color() +
  scale_color_manual(values = species_colors) +
scale_linetype_manual(values = c(
  "Significant"     = "solid",
  "Marginal"        = "dashed",
  "Not significant" = "dotted"
)) +
scale_alpha_manual(values = c(
  "Significant"     = 1.0,
  "Marginal"        = 0.7,
  "Not significant" = 0.35
)) +
  scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
) +
  labs(x = "Canopy closure (%)",
       y = expression(P[50]~"(MPa)"))   +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  
    panel.grid = element_blank(),      
    axis.line = element_blank(),      
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig2C

# ---- FIGURE 2D: P50_combined by elevation ----
all_preds_elev_P50 <- data.frame()
model_summaries_elev_P50 <- data.frame()

for (sp in names(species_list_elev)) {
  df <- species_list_elev[[sp]]

  # Fit model
  mod <- lmer(P50_combined ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot) + (1|year), data = df)

  # Extract fixed-effect p-value for elevation
  summ <- summary(mod)$coefficients
  elev_p <- summ["elev_indiv_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  elev_p < 0.05 ~ "Significant",
  elev_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)

  model_summaries_elev_P50 <- dplyr::bind_rows(model_summaries_elev_P50,
                                      data.frame(species = sp,
                                                 p_value = elev_p,
                                                 signif = signif_label))

  # Compute mean and SD for scaling elevation
  mean_elev <- mean(df$elev_indiv, na.rm = TRUE)
  sd_elev   <- sd(df$elev_indiv, na.rm = TRUE)

  # Sequence of elevations
  elev_seq <- seq(min(df$elev_indiv, na.rm = TRUE),
                  max(df$elev_indiv, na.rm = TRUE),
                  length.out = 100)

  # Prediction data
  newdata <- data.frame(
    elev_indiv_scaled = (elev_seq - mean_elev)/sd_elev,
    diameter_complete_scaled = 0,
    predawn_scaled = 0,
    canopy_closure_scaled = 0
  )

  newdata$elevation <- elev_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds_elev_P50 <- dplyr::bind_rows(all_preds_elev_P50, newdata)
}

Fig2D <- ggplot() +
  geom_point(data = trait_data1 %>% filter(species %in% names(species_list_elev)),
             aes(x = elev_indiv, y = P50_combined, color = species),
             alpha = 0.35, size = 2.5, na.rm = TRUE) +
  geom_line(data = all_preds_elev_P50,
            aes(x = elevation, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +  
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  scale_y_reverse(labels = function(x) paste0("\u2212", x)) +
  labs(x = "Elevation (ft)", y = expression(P[50]~"(MPa)")) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),
    panel.grid = element_blank(),
    axis.line = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )

grid.arrange(Fig2A, Fig2B, Fig2C, Fig2D, ncol=2, nrow=2)
```

#Figure 3 - HSM
## plotting model fit
```{r}

### FIGURE 3A: HSM_predawn_P50_mean
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  if (sp == "ABMA") {
    mod <- lmer(HSM_predawn_P50_mean ~ diameter_complete_scaled + canopy_closure_scaled + (1|plot) + (1|year), data = df)
  } else {
    mod <- lmer(HSM_predawn_P50_mean ~ diameter_complete_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot) + (1|year), data = df)
  }

  summ <- summary(mod)$coefficients
  diam_p <- summ["diameter_complete_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  diam_p < 0.05 ~ "Significant",
  diam_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)
  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = diam_p,
                                                 signif = signif_label))

  mean_d <- mean(df$diameter_complete, na.rm = TRUE)
  sd_d   <- sd(df$diameter_complete, na.rm = TRUE)

  diam_seq <- seq(min(df$diameter_complete, na.rm = TRUE),
                  max(df$diameter_complete, na.rm = TRUE),
                  length.out = 100)

  if (sp == "ABMA") {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_d)/sd_d,
      canopy_closure_scaled = 0
    )
  } else {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_d)/sd_d,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0
    )
  }

  newdata$diameter_cm <- diam_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

Fig3A <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = diameter_complete, y = HSM_predawn_P50_mean, color = species),
             alpha = 0.35, size = 2, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = diameter_cm, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  labs(x = "Basal diameter (cm)",
       y = expression(HSM[predawn]~"(MPa)")) + 
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  
    panel.grid = element_blank(),       
    axis.line = element_blank(),         
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig3A
### FIGURE 3B: HSM_midday_P50_mean
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  if (sp == "ABMA") {
    mod <- lmer(HSM_midday_P50_mean ~ diameter_complete_scaled + canopy_closure_scaled + (1|plot) + (1|year), data = df)
  } else {
    mod <- lmer(HSM_midday_P50_mean ~ diameter_complete_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot) + (1|year), data = df)
  }

  summ <- summary(mod)$coefficients
  diam_p <- summ["diameter_complete_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  diam_p < 0.05 ~ "Significant",
  diam_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)

  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = diam_p,
                                                 signif = signif_label))

  mean_d <- mean(df$diameter_complete, na.rm = TRUE)
  sd_d   <- sd(df$diameter_complete, na.rm = TRUE)

  diam_seq <- seq(min(df$diameter_complete, na.rm = TRUE),
                  max(df$diameter_complete, na.rm = TRUE),
                  length.out = 100)

  if (sp == "ABMA") {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_d)/sd_d,
      canopy_closure_scaled = 0
    )
  } else {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_d)/sd_d,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0
    )
  }

  newdata$diameter_cm <- diam_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

Fig3B <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = diameter_complete, y = HSM_midday_P50_mean, color = species),
             alpha = 0.35, size = 2, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = diameter_cm, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
   scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  labs(x = "Basal diameter (cm)",
       y = expression(HSM[midday]~"(MPa)")) + 
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  
    panel.grid = element_blank(),       
    axis.line = element_blank(),         
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig3B

# Unified y-axis
ymin <- min(c(trait_data1$HSM_predawn_P50_mean, trait_data1$HSM_midday_P50_mean), na.rm = TRUE)
ymax <- max(c(trait_data1$HSM_predawn_P50_mean, trait_data1$HSM_midday_P50_mean), na.rm = TRUE)

Fig3A <- Fig3A + coord_cartesian(ylim = c(ymin, ymax))
Fig3B <- Fig3B + coord_cartesian(ylim = c(ymin, ymax))

grid.arrange(Fig3A, Fig3B, nrow=1, ncol = 2)

### FIGURE 3C: HSM_predawn_P50 by percent cover
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(HSM_predawn_P50_mean ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled +
                  (1|plot) + (1|year), data = df)
  } else {
    mod <- lmer(HSM_predawn_P50_mean ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled +
                  elev_indiv_scaled + (1|plot) + (1|year), data = df)
  }

  # Extract fixed-effect p-value for canopy closure
  summ <- summary(mod)$coefficients
  cc_p <- summ["canopy_closure_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  cc_p < 0.05 ~ "Significant",
  cc_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)

  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = cc_p,
                                                 signif = signif_label))

  # Compute mean and SD for scaling percent cover
  mean_cc <- mean(df$percent_cover, na.rm = TRUE)
  sd_cc   <- sd(df$percent_cover, na.rm = TRUE)

  # Sequence of percent cover values
  cc_seq <- seq(min(df$percent_cover, na.rm = TRUE),
                max(df$percent_cover, na.rm = TRUE),
                length.out = 100)

  # Prediction data
  if (sp == "ABMA") {
    newdata <- data.frame(
      canopy_closure_scaled = (cc_seq - mean_cc)/sd_cc,
      diameter_complete_scaled = 0,
      predawn_scaled = 0
    )
  } else {
    newdata <- data.frame(
      canopy_closure_scaled = (cc_seq - mean_cc)/sd_cc,
      diameter_complete_scaled = 0,
      predawn_scaled = 0,
      elev_indiv_scaled = 0
    )
  }

  newdata$percent_cover <- cc_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### Plot
Fig3C <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = percent_cover, y = HSM_predawn_P50_mean, color = species),
             alpha = 0.35, size = 2, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = percent_cover, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
    scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  labs(x = "Canopy closure (%)",
       y = expression(HSM[predawn]~"(MPa)"))  + 
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  
    panel.grid = element_blank(),       
    axis.line = element_blank(),         
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig3C

### FIGURE 3D: HSM_midday_P50 by percent cover
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(HSM_midday_P50_mean ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled +
                  (1|plot) + (1|year), data = df)
  } else {
    mod <- lmer(HSM_midday_P50_mean ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled +
                  elev_indiv_scaled + (1|plot) + (1|year), data = df)
  }

  # Extract fixed-effect p-value for canopy closure
  summ <- summary(mod)$coefficients
  cc_p <- summ["canopy_closure_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  cc_p < 0.05 ~ "Significant",
  cc_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)

  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = cc_p,
                                                 signif = signif_label))

  # Compute mean and SD for scaling percent cover
  mean_cc <- mean(df$percent_cover, na.rm = TRUE)
  sd_cc   <- sd(df$percent_cover, na.rm = TRUE)

  # Sequence of percent cover values
  cc_seq <- seq(min(df$percent_cover, na.rm = TRUE),
                max(df$percent_cover, na.rm = TRUE),
                length.out = 100)

  # Prediction data
  if (sp == "ABMA") {
    newdata <- data.frame(
      canopy_closure_scaled = (cc_seq - mean_cc)/sd_cc,
      diameter_complete_scaled = 0,
      predawn_scaled = 0
    )
  } else {
    newdata <- data.frame(
      canopy_closure_scaled = (cc_seq - mean_cc)/sd_cc,
      diameter_complete_scaled = 0,
      predawn_scaled = 0,
      elev_indiv_scaled = 0
    )
  }

  newdata$percent_cover <- cc_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### Plot
Fig3D <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = percent_cover, y = HSM_midday_P50_mean, color = species),
             alpha = 0.35, size = 2, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = percent_cover, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
   scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  labs(x = "Canopy closure (%)",
       y = expression(HSM[midday]~"(MPa)"))+ 
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  
    panel.grid = element_blank(),       
    axis.line = element_blank(),         
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig3D

# ---- FIGURE 3E: HSM_predawn_P50_mean by elevation ----
all_preds_elev_HSM <- data.frame()
model_summaries_elev_HSM <- data.frame()

for (sp in names(species_list_elev)) {
  df <- species_list_elev[[sp]]

  # Fit model
  mod <- lmer(HSM_predawn_P50_mean ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot) + (1|year), data = df)

  # Extract fixed-effect p-value for elevation
  summ <- summary(mod)$coefficients
  elev_p <- summ["elev_indiv_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  elev_p < 0.05 ~ "Significant",
  elev_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)

  model_summaries_elev_HSM <- dplyr::bind_rows(model_summaries_elev_HSM,
                                      data.frame(species = sp,
                                                 p_value = elev_p,
                                                 signif = signif_label))

  # Compute mean and SD for scaling elevation
  mean_elev <- mean(df$elev_indiv, na.rm = TRUE)
  sd_elev   <- sd(df$elev_indiv, na.rm = TRUE)

  # Sequence of elevations
  elev_seq <- seq(min(df$elev_indiv, na.rm = TRUE),
                  max(df$elev_indiv, na.rm = TRUE),
                  length.out = 100)

  # Prediction data
  newdata <- data.frame(
    elev_indiv_scaled = (elev_seq - mean_elev)/sd_elev,
    diameter_complete_scaled = 0,
    predawn_scaled = 0,
    canopy_closure_scaled = 0
  )

  newdata$elevation <- elev_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds_elev_HSM <- dplyr::bind_rows(all_preds_elev_HSM, newdata)
}

# Plot Figure 3E
Fig3E <- ggplot() +
  geom_point(data = trait_data1 %>% dplyr::filter(species %in% names(species_list_elev)),
             aes(x = elev_indiv, y = HSM_predawn_P50_mean, color = species),
             alpha = 0.35, size = 2, na.rm = TRUE) +
  geom_line(data = all_preds_elev_HSM,
            aes(x = elevation, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  labs(x = "Elevation (ft)",
       y = expression(HSM[predawn]~"(MPa)"))+ 
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  
    panel.grid = element_blank(),       
    axis.line = element_blank(),         
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig3E

# ---- FIGURE 3F: HSM_midday_P50_mean by elevation ----
all_preds_elev_HSM <- data.frame()
model_summaries_elev_HSM <- data.frame()

for (sp in names(species_list_elev)) {
  df <- species_list_elev[[sp]]

  # Fit model
  mod <- lmer(HSM_midday_P50_mean ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot) + (1|year), data = df)

  # Extract fixed-effect p-value for elevation
  summ <- summary(mod)$coefficients
  elev_p <- summ["elev_indiv_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  elev_p < 0.05 ~ "Significant",
  elev_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)

  model_summaries_elev_HSM <- dplyr::bind_rows(model_summaries_elev_HSM,
                                      data.frame(species = sp,
                                                 p_value = elev_p,
                                                 signif = signif_label))

  # Compute mean and SD for scaling elevation
  mean_elev <- mean(df$elev_indiv, na.rm = TRUE)
  sd_elev   <- sd(df$elev_indiv, na.rm = TRUE)

  # Sequence of elevations
  elev_seq <- seq(min(df$elev_indiv, na.rm = TRUE),
                  max(df$elev_indiv, na.rm = TRUE),
                  length.out = 100)

  # Prediction data
  newdata <- data.frame(
    elev_indiv_scaled = (elev_seq - mean_elev)/sd_elev,
    diameter_complete_scaled = 0,
    predawn_scaled = 0,
    canopy_closure_scaled = 0
  )

  newdata$elevation <- elev_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds_elev_HSM <- dplyr::bind_rows(all_preds_elev_HSM, newdata)
}

# Plot Figure 3F
Fig3F <- ggplot() +
  geom_point(data = trait_data1 %>% dplyr::filter(species %in% names(species_list_elev)),
             aes(x = elev_indiv, y = HSM_midday_P50_mean, color = species),
             alpha = 0.35, size = 2, na.rm = TRUE) +
  geom_line(data = all_preds_elev_HSM,
            aes(x = elevation, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  labs(x = "Elevation (ft)",
       y = expression(HSM[midday]~"(MPa)")) + 
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  
    panel.grid = element_blank(),       
    axis.line = element_blank(),         
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig3F

### FIGURE 3G: HSM by predawn 
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(HSM_midday_P50_mean ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled + (1|plot) + (1|year), data = df)
  } else {
    mod <- lmer(HSM_midday_P50_mean ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot) + (1|year), data = df)
  }

  # Extract fixed-effect p-value for predawn
  summ <- summary(mod)$coefficients
  pd_p <- summ["predawn_scaled", "Pr(>|t|)"]
  signif_label <- ifelse(pd_p < 0.05, "Significant", "Not significant")

  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = pd_p,
                                                 signif = signif_label))

  # Compute mean and SD for scaling predawn
  mean_pd <- mean(df$predawn_MPa_combined, na.rm = TRUE)
  sd_pd   <- sd(df$predawn_MPa_combined, na.rm = TRUE)

  # Sequence of predawns
  pd_seq <- seq(min(df$predawn_MPa_combined, na.rm = TRUE),
                  max(df$predawn_MPa_combined, na.rm = TRUE),
                  length.out = 100)

  # Prediction data
  if (sp == "ABMA") {
    newdata <- data.frame(
      predawn_scaled = (pd_seq - mean_pd)/sd_pd,
      diameter_complete_scaled = 0,
      canopy_closure_scaled = 0
    )
  } else {
    newdata <- data.frame(
      predawn_scaled = (pd_seq - mean_pd)/sd_pd,
      diameter_complete_scaled = 0,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0
    )
  }

  newdata$predawn <- pd_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### FIGURE 3G: HSM by predawn
Fig3G <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = predawn_MPa_combined, y = HSM_midday_P50_mean, color = species),
             alpha = 0.35, size = 2, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = predawn, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  scale_x_reverse(
    labels = function(x) paste0("\u2212", abs(x))
  ) +
  labs(x = expression(Psi[predawn]~"(MPa)"),
       y = expression(HSM[midday]~"(MPa)")) + 
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),  
    panel.grid = element_blank(),       
    axis.line = element_blank(),         
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )

Fig3G

blank_plot <- ggplot() + theme_void()

grid.arrange(Fig3A, Fig3B, Fig3C, Fig3D, Fig3E, Fig3F, blank_plot, Fig3G, ncol =2, nrow = 4)
#Figure3 <- list(Fig3A, Fig3B, Fig3C, Fig3D, Fig3E, Fig3F, Fig3G)

#Figure3_legend <- cowplot::get_legend(
 # Figure3[[3]] +
  #  theme(legend.position = "right",
   #       legend.title = element_blank())
#)

#grid.arrange(
 # Figure3[[1]], Figure3[[2]],
  #Figure3[[3]], Figure3[[4]],
 # Figure3[[5]], Figure3_legend,
  #ncol = 2,
 # nrow = 3
#)


#grid.arrange(Fig3A, Fig3B, Fig3C, Fig3D, Fig3E, nrow = 3, ncol =2)

```


# Figure 4 - growth 

```{r}

g1 <- ggplot(trait_data1, aes(x = log(diameter_complete), y = log(internode_length_1_2), color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  labs(x = "log Diameter at root collar (cm)", y = "log Internode length (cm)") +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse()  +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )


g1 

g2 <- ggplot(trait_data1, aes(x = predawn_MPa_combined, y = internode_length_1_2, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  labs(x = "Predawn water potential (-MPa)", y = "Internode length (cm)") +
  xlim(NA, 1.7) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse()  +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )


g2


g3 <- ggplot(trait_data1, aes(x =HSM_midday_P50_mean , y = internode_length_1_2, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +  
  labs(
    x = "HSM midday (MPa)",
    y = "Internode length (cm) ",
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse() + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )


g3

g4 <- ggplot(trait_data1, aes(x = P50_combined, y = internode_length_1_2, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(y = "Internode length (cm)",
       x = expression(P[50]~"(-MPa)")) + 
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse() + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

g4

g5 <- ggplot(trait_data1, aes(x = percent_cover, y = internode_length_1_2, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  labs(x = "Canopy closure (%)", y = "Internode length (cm)") +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse() + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

g5

g6 <- trait_data1 %>%
  filter(species != "ABMA") %>%
  ggplot(aes(x = elev_indiv, y = internode_length_1_2, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  labs(x = "Elevation (ft)", y = "Internode length (cm)") +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse() + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )


g6
grid.arrange(g1, g2, g3, g4, g5,g6, nrow = 3, ncol = 2)

```

## plotting model fit
```{r}
### Figure 4A: Internode length by diameter, holding predawn, canopy closure, and elevation constant

all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]
  
  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(internode_length_1_2 ~ diameter_complete_scaled + predawn_scaled +
                  canopy_closure_scaled + (1|plot),
                data = df)
  } else {
    mod <- lmer(internode_length_1_2 ~ diameter_complete_scaled + predawn_scaled +
                  canopy_closure_scaled + elev_indiv_scaled + (1|plot),
                data = df)
  }
  
  # Extract fixed-effect p-value for diameter
  summ <- summary(mod)$coefficients
  if ("diameter_complete_scaled" %in% rownames(summ)) {
    diam_p <- summ["diameter_complete_scaled", "Pr(>|t|)"]
  } else {
    diam_p <- NA
  }
   signif_label <- dplyr::case_when(
  diam_p < 0.05 ~ "Significant",
  diam_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)
  
  # Sequence of diameter values 
  mean_diam <- mean(df$diameter_complete, na.rm = TRUE)
  sd_diam <- sd(df$diameter_complete, na.rm = TRUE)
  diam_seq <- seq(min(df$diameter_complete, na.rm = TRUE),
                  max(df$diameter_complete, na.rm = TRUE),
                  length.out = 100)
  
  # Create new data with scaled diameter varying and others held constant
  if (sp == "ABMA") {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_diam) / sd_diam,
      predawn_scaled = 0,
      canopy_closure_scaled = 0
    )
  } else {
    newdata <- data.frame(
      diameter_complete_scaled = (diam_seq - mean_diam) / sd_diam,
      predawn_scaled = 0,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0
    )
  }
  
  # Predictions
  newdata$diameter_complete <- diam_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label
  
  all_preds <- bind_rows(all_preds, newdata)
}

### Figure 4A
Fig4A <- ggplot() +
  geom_point(data = bind_rows(species_list),
             aes(x = diameter_complete, y = internode_length_1_2, color = species),
             alpha = 0.35, size = 2.5, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = diameter_complete, y = pred, color = species, linetype = signif,
      alpha = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  labs(x = "Basal diameter (cm)", y = "Internode length (cm)")  +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),
    panel.grid = element_blank(),
    axis.line = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig4A

### FIGURE 4B: Internode length by predawn water potential
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(internode_length_1_2 ~ predawn_scaled + canopy_closure_scaled + diameter_complete_scaled +
                  (1|plot), data = df)
  } else {
    mod <- lmer(internode_length_1_2 ~ predawn_scaled + canopy_closure_scaled + elev_indiv_scaled + diameter_complete_scaled +
                  (1|plot), data = df)
  }

  # Extract fixed-effect p-value for predawn
  summ <- summary(mod)$coefficients
  pd_p <- summ["predawn_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  pd_p < 0.05 ~ "Significant",
  pd_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)
  # Prediction data
  mean_pd <- mean(df$predawn_MPa_combined, na.rm = TRUE)
  sd_pd   <- sd(df$predawn_MPa_combined, na.rm = TRUE)
  pd_seq <- seq(min(df$predawn_MPa_combined, na.rm = TRUE),
                max(df$predawn_MPa_combined, na.rm = TRUE),
                length.out = 100)

  if (sp == "ABMA") {
    newdata <- data.frame(
      predawn_scaled = (pd_seq - mean_pd)/sd_pd,
      canopy_closure_scaled = 0,
      diameter_complete_scaled = 0
    )
  } else {
    newdata <- data.frame(
      predawn_scaled = (pd_seq - mean_pd)/sd_pd,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0,
      diameter_complete_scaled = 0
    )
  }

  newdata$predawn <- pd_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### Plot Figure 4B
Fig4B <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = predawn_MPa_combined, y = internode_length_1_2, color = species),
             alpha = 0.35, size = 2.5, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = predawn, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  ))  +
  scale_x_reverse(
    labels = function(x) paste0("\u2212", abs(x))
  ) +
  labs(x = expression(Psi[predawn]~"(MPa)"),
       y = "Internode length (cm)")  +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),
    panel.grid = element_blank(),
    axis.line = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig4B

### FIGURE 4C: Internode length by HSM_midday_P50_mean
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(internode_length_1_2 ~ HSM_midday_P50_mean_scaled + predawn_scaled + canopy_closure_scaled + diameter_complete_scaled +
                  (1|plot), data = df)
  } else {
    mod <- lmer(internode_length_1_2 ~ HSM_midday_P50_mean_scaled + predawn_scaled + canopy_closure_scaled + elev_indiv_scaled + diameter_complete_scaled +
                  (1|plot), data = df)
  }

  # Extract fixed-effect p-value for HSM_midday_P50_mean
  summ <- summary(mod)$coefficients
  hsm_p <- summ["HSM_midday_P50_mean_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  hsm_p < 0.05 ~ "Significant",
  hsm_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)

  # Prediction data
  mean_hsm <- mean(df$HSM_midday_P50_mean, na.rm = TRUE)
  sd_hsm   <- sd(df$HSM_midday_P50_mean, na.rm = TRUE)
  hsm_seq <- seq(min(df$HSM_midday_P50_mean, na.rm = TRUE),
                 max(df$HSM_midday_P50_mean, na.rm = TRUE),
                 length.out = 100)

  if (sp == "ABMA") {
    newdata <- data.frame(
      HSM_midday_P50_mean_scaled = (hsm_seq - mean_hsm)/sd_hsm,
      predawn_scaled = 0,
      canopy_closure_scaled = 0,
      diameter_complete_scaled = 0
    )
  } else {
    newdata <- data.frame(
      HSM_midday_P50_mean_scaled = (hsm_seq - mean_hsm)/sd_hsm,
      predawn_scaled = 0,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0,
      diameter_complete_scaled = 0
    )
  }

  newdata$HSM_midday_P50_mean <- hsm_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### Plot Figure 4C
Fig4C <- ggplot() +
  geom_point(data = trait_data1,
             aes(x = HSM_midday_P50_mean, y = internode_length_1_2, color = species),
             alpha = 0.35, size = 2.5, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = HSM_midday_P50_mean, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  labs(x = expression(HSM[midday]~"(MPa)"),
       y = "Internode length (cm)") +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),
    panel.grid = element_blank(),
    axis.line = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig4C
# Figure 4D
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]

  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(internode_length_1_2 ~ diameter_complete_scaled + P50_scaled + predawn_scaled + canopy_closure_scaled +
                  (1|plot), data = df)
  } else {
    mod <- lmer(internode_length_1_2 ~ diameter_complete_scaled + P50_scaled + predawn_scaled + canopy_closure_scaled + elev_indiv_scaled +
                  (1|plot), data = df)
  }

  # Extract fixed-effect p-value for P50_scaled
  summ <- summary(mod)$coefficients
  p50_p <- summ["P50_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  p50_p < 0.05 ~ "Significant",
  p50_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)
  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = p50_p,
                                                 signif = signif_label))

  # Prediction data
  mean_P50 <- mean(df$P50_combined, na.rm = TRUE)
  sd_P50 <- sd(df$P50_combined, na.rm = TRUE)
  P50_seq <- seq(min(df$P50_combined, na.rm = TRUE), max(df$P50_combined, na.rm = TRUE), length.out = 100)

  if (sp == "ABMA") {
    newdata <- data.frame(
      P50_scaled = (P50_seq - mean_P50) / sd_P50,
      diameter_complete_scaled = 0,
      predawn_scaled = 0,
      canopy_closure_scaled = 0
    )
  } else {
    newdata <- data.frame(
      P50_scaled = (P50_seq - mean_P50) / sd_P50,
      diameter_complete_scaled = 0,
      predawn_scaled = 0,
      canopy_closure_scaled = 0,
      elev_indiv_scaled = 0
    )
  }
  newdata$P50_raw <- P50_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### FIGURE 4D: Internode length by P50
Fig4D <- ggplot() +
  geom_point(data = bind_rows(species_list), 
             aes(x = P50_combined, y = internode_length_1_2, color = species),
             alpha = 0.35, size = 2.5, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = P50_raw, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  scale_x_reverse(
    labels = function(x) paste0("\u2212", abs(x))
  ) +
  labs(x = expression(P[50]~"(MPa)"), y = "Internode length (cm)")  +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),
    panel.grid = element_blank(),
    axis.line = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig4D

### FIGURE 4E: Internode length by canopy closure with predawn as covariate
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list)) {
  df <- species_list[[sp]]
  
  # Fit species-specific model
  if (sp == "ABMA") {
    mod <- lmer(internode_length_1_2 ~ diameter_complete_scaled + canopy_closure_scaled + predawn_scaled + (1|plot), data = df)
  } else {
    mod <- lmer(internode_length_1_2 ~ diameter_complete_scaled + canopy_closure_scaled + predawn_scaled + elev_indiv_scaled + (1|plot), data = df)
  }
  
  # Extract fixed-effect p-value for canopy closure
  summ <- summary(mod)$coefficients
  canopy_p <- summ["canopy_closure_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  canopy_p < 0.05 ~ "Significant",
  canopy_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)
  
  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = canopy_p,
                                                 signif = signif_label))
  
  # Compute mean and SD for scaling canopy closure
  mean_canopy <- mean(df$percent_cover, na.rm = TRUE)
  sd_canopy   <- sd(df$percent_cover, na.rm = TRUE)
  
  # Sequence of canopy closure values
  canopy_seq <- seq(min(df$percent_cover, na.rm = TRUE),
                    max(df$percent_cover, na.rm = TRUE),
                    length.out = 100)
  
  # Prediction data
  if (sp == "ABMA") {
    newdata <- data.frame(
      diameter_complete_scaled = 0,
      predawn_scaled = 0,
      canopy_closure_scaled = (canopy_seq - mean_canopy)/sd_canopy
    )
  } else {
    newdata <- data.frame(
      diameter_complete_scaled = 0,
      predawn_scaled = 0,
      canopy_closure_scaled = (canopy_seq - mean_canopy)/sd_canopy,
      elev_indiv_scaled = 0
    )
  }
  
  newdata$canopy <- canopy_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label
  
  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### FIGURE 4E: Internode length by canopy closure
Fig4E <- ggplot() +
  geom_point(data = bind_rows(species_list),
             aes(x = percent_cover, y = internode_length_1_2, color = species),
             alpha = 0.25, size = 2.5, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = canopy, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  labs(x = "Canopy closure (%)",
       y = "Internode length (cm)")  +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),
    panel.grid = element_blank(),
    axis.line = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
Fig4E
### FIGURE 4F: Internode length by elevation
all_preds <- data.frame()
model_summaries <- data.frame()

for (sp in names(species_list_elev)) {
  df <- species_list_elev[[sp]]

  # Fit species-specific model
  mod <- lmer(internode_length_1_2 ~ diameter_complete_scaled + predawn_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot), data = df)

  # Extract fixed-effect p-value for elevation
  summ <- summary(mod)$coefficients
  elev_p <- summ["elev_indiv_scaled", "Pr(>|t|)"]
  signif_label <- dplyr::case_when(
  elev_p < 0.05 ~ "Significant",
  elev_p < 0.1  ~ "Marginal",
  TRUE          ~ "Not significant"
)

  model_summaries <- dplyr::bind_rows(model_summaries,
                                      data.frame(species = sp,
                                                 p_value = elev_p,
                                                 signif = signif_label))

  # Sequence of elevation values
  elev_seq <- seq(min(df$elev_indiv, na.rm = TRUE),
                  max(df$elev_indiv, na.rm = TRUE),
                  length.out = 100)
  # Prediction data
  newdata <- data.frame(
    diameter_complete_scaled = 0,
    predawn_scaled = 0,
    canopy_closure_scaled = 0,
    elev_indiv_scaled = (elev_seq - mean(df$elev_indiv, na.rm = TRUE))/sd(df$elev_indiv, na.rm = TRUE)
  )

  newdata$elev <- elev_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds <- dplyr::bind_rows(all_preds, newdata)
}

### FIGURE 4F plot
Fig4F <- ggplot() +
  geom_point(data = bind_rows(species_list_elev),
             aes(x = elev_indiv, y = internode_length_1_2, color = species),
             alpha = 0.25, size = 2.5, na.rm = TRUE) +
  geom_line(data = all_preds,
            aes(x = elev, y = pred, color = species, linetype = signif),
            size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_alpha_manual(values = c(
    "Significant"     = 1.0,
    "Marginal"        = 0.7,
    "Not significant" = 0.35
  )) +
  labs(x = "Elevation (ft)",
       y = "Internode length (cm)")  +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray50", fill = NA, size = 1),
    panel.grid = element_blank(),
    axis.line = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )

Fig4F
grid.arrange(Fig4A, Fig4B, Fig4C, Fig4D, Fig4E, Fig4F, ncol = 2, nrow =3)

```

# Figure 4 - growth~diam
```{r}
# if growth is size dependent, lets standardize growth by diameter

# Fit linear model of growth on diameter
growth_diam_mod <- lm(internode_length_1_2 ~ diameter_complete, data = trait_data1, na.action = na.exclude)

# Extract residuals and add as new column
trait_data1$growth_residuals <- residuals(growth_diam_mod)

# subset dataframe by species 
ABCO <- trait_data1[trait_data1$species == 'ABCO',]
PILA <- trait_data1[trait_data1$species == 'PILA',]
PIPO <- trait_data1[trait_data1$species == 'PIPO',]
ABMA <- trait_data1[trait_data1$species == 'ABMA',]


g2 <- ggplot(trait_data1, aes(x = -predawn_MPa_combined, y = growth_residuals, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(
    x = "Predawn water potential (-MPa)",
    y = "growth~diam residuals"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
g2

g3 <- ggplot(trait_data1, aes(x = HSM_midday_P50_mean, y = growth_residuals, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +  
  labs(
    x = "HSM midday (-MPa)",
    y = "growth~diam residuals",
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise", "PIPO" = "lawngreen",
    "ABMA" = "maroon3", "PILA" = "lightslateblue"
  ))  +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )
g3

g4 <- ggplot(trait_data1, aes(x = P50_combined, y = growth_residuals, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  labs(x = expression(P[50]~"(-MPa)", y = "growth~diam residuals") +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise", "PIPO" = "lawngreen",
    "ABMA" = "maroon3", "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )
g4 

g5 <- ggplot(trait_data1, aes(x = percent_cover, y = growth_residuals, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  labs(x = "Canopy closure (%)", y = "growth~diam residuals") +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise", "PIPO" = "lawngreen",
    "ABMA" = "maroon3", "PILA" = "lightslateblue"
  ))  +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),         # remove full box
    axis.line = element_line(color = "black"),  # keep x and y axes
    legend.position = "none"
  )
g5

g6 <- trait_data1 %>%
  filter(species != "ABMA") %>%
  ggplot(aes(x = elev_indiv, y = growth_residuals, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(x = "Elevation (ft)", y = "growth~diam residuals") +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "none"
  )

g6
grid.arrange(g1, g2, g3, g4, g5,g6, nrow = 3, ncol = 2)
```


```{r}
library(ggplot2)

ggplot(trait_data2, aes(x = DBH_complete, y = water_potential)) +
  geom_point(aes(shape = type, color = HSM_predawn_P50_mean), size = 3) +
  scale_shape_manual(values = c("predawn_MPa" = 16, "P50_mean" = 17, "P50_MPa" = 17)) +  # circle, triangle
  scale_color_gradient(low = "red", high = "blue", name = "HSM\n(P50 - predawn)") +
  facet_wrap(~species, ncol = 1) +
  theme_bw() +
  labs(
    x = "Diameter at root collar (cm)",
    y = "Water Potential (-MPa)",
    shape = "Type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )



library(dplyr)
library(tidyr)

# Create wide format with HSM
HSM_color_gradient <- trait_data2 %>%
  filter(type %in% c("predawn_MPa", "P50_mean")) %>%
  select(tree_ID, species, DBH_complete, type, water_potential) %>%
  pivot_wider(names_from = type, values_from = water_potential) %>%
  mutate(HSM = P50_mean - predawn_MPa)


library(ggplot2)

ggplot() +
  # Colored connecting lines
  geom_segment(data = HSM_color_gradient,
               aes(x = DBH_complete, xend = DBH_complete,
                   y = predawn_MPa, yend = P50_mean, color = HSM),
               size = 1) +
  
  # Points for both predawn and P50
  geom_point(data = HSM_color_gradient,
             aes(x = DBH_complete, y = predawn_MPa), 
             color = "steelblue", size = 3) +
  
  geom_point(data = HSM_color_gradient,
             aes(x = DBH_complete, y = P50_mean), 
             color = "darkblue", size = 3, shape = 17) +
  
  scale_color_gradient(low = "red", high = "blue", name = "HSM\n(P50 - predawn)") +
  facet_wrap(~species, ncol = 1) +
  theme_bw() +
  labs(
    x = "Diameter at root collar (cm)",
    y = "Water Potential (-MPa)"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


# Build polygons: x stays same, y goes from predawn to P50, loop back
HSM_polygons <- HSM_color_gradient %>%
  rowwise() %>%
  mutate(
    x1 = DBH_complete - 0.2,  # small width to create polygon
    x2 = DBH_complete + 0.2,
    y1 = predawn_MPa,
    y2 = P50_mean
  ) %>%
  ungroup() %>%
  rowwise() %>%
  do(data.frame(
    tree_ID = .$tree_ID,
    species = .$species,
    HSM = .$HSM,
    x = c(.$x1, .$x2, .$x2, .$x1),
    y = c(.$y1, .$y1, .$y2, .$y2)
  )) %>%
  ungroup()


ggplot() +
  geom_polygon(data = HSM_polygons, 
               aes(x = x, y = y, group = tree_ID, fill = HSM),
               alpha = 0.5, color = NA) +

  geom_point(data = HSM_color_gradient, 
             aes(x = DBH_complete, y = predawn_MPa), 
             color = "steelblue", size = 3) +

  geom_point(data = HSM_color_gradient, 
             aes(x = DBH_complete, y = P50_mean), 
             color = "darkblue", shape = 17, size = 3) +

  scale_fill_gradient(low = "red", high = "blue", name = "HSM\n(P50 - predawn)") +

  facet_wrap(~species, ncol = 1) +
  theme_bw() +
  labs(
    x = "Diameter at root collar (cm)",
    y = "Water Potential (-MPa)"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )



```
# Supplemental Figures
```{r}
S1A <- ggplot(subset(trait_data1, species != "ABMA"), aes(x = elev_indiv, y = predawn_MPa_2024, color = species))+
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "Predawn WP 2024 (-MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

S1A

S1B <- ggplot(subset(trait_data1, species != "ABMA"), aes(x = elev_indiv, y = midday_MPa_2024, color = species))+
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "Midday WP 2024 (-MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

S1B

S1C <- ggplot(subset(trait_data1, species != "ABMA"), aes(x = elev_indiv, y = predawn_MPa_2025, color = species))+
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "Predawn WP 2025 (-MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

S1C

S1D <- ggplot(subset(trait_data1, species != "ABMA"), aes(x = elev_indiv, y = midday_MPa_2025, color = species))+
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "Midday WP (2025) (-MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

S1D

S1E <- ggplot(subset(trait_data1, species != "ABMA"), aes(x = elev_indiv, y = predawn_MPa_combined, color = species))+
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "Predawn WP (2-year average) (-MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")

S1E

S1F <- ggplot(subset(trait_data1, species != "ABMA"), aes(x = elev_indiv, y = midday_MPa_combined, color = species))+
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "Midday WP (2-year average) (-MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

S1F

grid.arrange(S1A, S1B, S1C, S1D, S1E, S1F, nrow =3, ncol=2)
```
## plotting model fit

```{r}
### FIGURE S1A: Predawn by elevation (model-based)

all_preds_elev <- data.frame()
model_summaries_elev <- data.frame()

for (sp in setdiff(names(species_list), "ABMA")) {

  df <- species_list[[sp]]

  # Fit species-specific model (same structure, elevation is focal)
  mod <- lmer(
    predawn_MPa_2024 ~ elev_indiv_scaled + diameter_complete_scaled +
      canopy_closure_scaled + (1|plot),
    data = df
  )

  # Extract p-value for elevation
  summ <- summary(mod)$coefficients
  elev_p <- summ["elev_indiv_scaled", "Pr(>|t|)"]

  signif_label <- dplyr::case_when(
    elev_p < 0.05 ~ "Significant",
    elev_p < 0.1  ~ "Marginal",
    TRUE          ~ "Not significant"
  )

  model_summaries_elev <- dplyr::bind_rows(
    model_summaries_elev,
    data.frame(species = sp,
               p_value = elev_p,
               signif = signif_label)
  )

  # Compute mean & SD for elevation (unscaled)
  mean_e <- mean(df$elev_indiv, na.rm = TRUE)
  sd_e   <- sd(df$elev_indiv, na.rm = TRUE)

  # Elevation sequence
  elev_seq <- seq(min(df$elev_indiv, na.rm = TRUE),
                  max(df$elev_indiv, na.rm = TRUE),
                  length.out = 100)

  # Prediction data (other covariates at mean = 0 scaled)
  newdata <- data.frame(
    elev_indiv_scaled = (elev_seq - mean_e) / sd_e,
    diameter_complete_scaled = 0,
    canopy_closure_scaled = 0
  )

  newdata$elev_ft <- elev_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds_elev <- dplyr::bind_rows(all_preds_elev, newdata)
}


S1A <- ggplot() +
  geom_point(
    data = subset(trait_data1, species != "ABMA"),
    aes(x = elev_indiv, y = predawn_MPa_2024, color = species),
    size = 3,
    alpha = 0.35,
    na.rm = TRUE
  ) +
  scale_color_manual(values = species_colors) +
  new_scale_color() +
  geom_line(
    data = all_preds_elev,
    aes(x = elev_ft, y = pred,
        color = species,
        linetype = signif),
    size = 1.2,
    na.rm = TRUE
  ) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant"     = "solid",
    "Marginal"        = "dashed",
    "Not significant" = "dotted"
  )) +
  scale_x_continuous(
    breaks = seq(
      floor(min(trait_data1$elev_indiv, na.rm = TRUE)),
      ceiling(max(trait_data1$elev_indiv, na.rm = TRUE)),
      by = 500
    )
  ) +
  labs(
    x = "Elevation (ft)",
    y = expression(Psi[predawn]~"(MPa)"~"2024")
  ) +
  scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
)+
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

S1A

# ======================
# S1B: model + predictions
# ======================

all_preds_S1B <- data.frame()
model_summaries_S1B <- data.frame()

for (sp in setdiff(names(species_list), "ABMA")) {

  df <- species_list[[sp]]

  mod <- lmer(
    midday_MPa_2024 ~ elev_indiv_scaled + diameter_complete_scaled +
      canopy_closure_scaled + (1|plot),
    data = df
  )

  summ <- summary(mod)$coefficients
  elev_p <- summ["elev_indiv_scaled", "Pr(>|t|)"]

  signif_label <- dplyr::case_when(
    elev_p < 0.05 ~ "Significant",
    elev_p < 0.1  ~ "Marginal",
    TRUE          ~ "Not significant"
  )

  mean_e <- mean(df$elev_indiv, na.rm = TRUE)
  sd_e   <- sd(df$elev_indiv, na.rm = TRUE)

  elev_seq <- seq(min(df$elev_indiv, na.rm = TRUE),
                  max(df$elev_indiv, na.rm = TRUE),
                  length.out = 100)

  newdata <- data.frame(
    elev_indiv_scaled = (elev_seq - mean_e) / sd_e,
    diameter_complete_scaled = 0,
    canopy_closure_scaled = 0
  )

  newdata$elev_ft <- elev_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds_S1B <- dplyr::bind_rows(all_preds_S1B, newdata)
}

# ======================
# S1B: plot
# ======================

S1B <- ggplot() +
  geom_point(
    data = subset(trait_data1, species != "ABMA"),
    aes(x = elev_indiv, y = midday_MPa_2024, color = species),
    size = 3, alpha = 0.35
  ) +
  scale_color_manual(values = species_colors) +
  new_scale_color() +
  geom_line(
    data = all_preds_S1B,
    aes(x = elev_ft, y = pred, color = species, linetype = signif),
    size = 1.2
  ) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant" = "solid",
    "Marginal" = "dashed",
    "Not significant" = "dotted"
  )) +
  labs(
    x = "Elevation (ft)",
    y = expression(Psi[midday]~"(MPa)"~"2024")
  ) +scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
)+
  scale_x_continuous(
    breaks = seq(
      floor(min(trait_data1$elev_indiv, na.rm = TRUE)),
      ceiling(max(trait_data1$elev_indiv, na.rm = TRUE)),
      by = 500
    )
  ) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.position = "none")

S1B


# ======================
# S1C: model + predictions
# ======================

all_preds_S1C <- data.frame()
model_summaries_S1C <- data.frame()

for (sp in setdiff(names(species_list), "ABMA")) {

  df <- species_list[[sp]]

  mod <- lmer(
    predawn_MPa_2025 ~ elev_indiv_scaled + diameter_complete_scaled +
      canopy_closure_scaled + (1 | plot),
    data = df
  )

  summ <- summary(mod)$coefficients
  elev_p <- summ["elev_indiv_scaled", "Pr(>|t|)"]

  signif_label <- dplyr::case_when(
    elev_p < 0.05 ~ "Significant",
    elev_p < 0.1 ~ "Marginal",
    TRUE ~ "Not significant"
  )

  model_summaries_S1C <- dplyr::bind_rows(
    model_summaries_S1C,
    data.frame(species = sp, p_value = elev_p, signif = signif_label)
  )

  mean_e <- mean(df$elev_indiv, na.rm = TRUE)
  sd_e <- sd(df$elev_indiv, na.rm = TRUE)

  elev_seq <- seq(
    min(df$elev_indiv, na.rm = TRUE),
    max(df$elev_indiv, na.rm = TRUE),
    length.out = 100
  )

  newdata <- data.frame(
    elev_indiv_scaled = (elev_seq - mean_e) / sd_e,
    diameter_complete_scaled = 0,
    canopy_closure_scaled = 0
  )

  newdata$elev_ft <- elev_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds_S1C <- dplyr::bind_rows(all_preds_S1C, newdata)
}

# ======================
# S1C: plot
# ======================

S1C <- ggplot() +
  geom_point(
    data = subset(trait_data1, species != "ABMA"),
    aes(x = elev_indiv, y = predawn_MPa_2025, color = species),
    size = 3,
    alpha = 0.35
  ) +
  scale_color_manual(values = species_colors) +
  ggnewscale::new_scale_color() +
  geom_line(
    data = all_preds_S1C,
    aes(x = elev_ft, y = pred, color = species, linetype = signif),
    size = 1.2
  ) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant" = "solid",
    "Marginal" = "dashed",
    "Not significant" = "dotted"
  )) +
  labs(
    x = "Elevation (ft)",
    y = expression(Psi[predawn]~"(MPa)"~"2025")
  ) +
  scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
)+
  scale_x_continuous(
    breaks = seq(
      floor(min(trait_data1$elev_indiv, na.rm = TRUE)),
      ceiling(max(trait_data1$elev_indiv, na.rm = TRUE)),
      by = 500
    )
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    legend.position = "none"
  )

S1C


# ======================
# S1D: model + predictions
# ======================

all_preds_S1D <- data.frame()
model_summaries_S1D <- data.frame()

for (sp in setdiff(names(species_list), "ABMA")) {

  df <- species_list[[sp]]

  mod <- lmer(
    midday_MPa_2025 ~ elev_indiv_scaled + diameter_complete_scaled +
      canopy_closure_scaled + (1 | plot),
    data = df
  )

  summ <- summary(mod)$coefficients
  elev_p <- summ["elev_indiv_scaled", "Pr(>|t|)"]

  signif_label <- dplyr::case_when(
    elev_p < 0.05 ~ "Significant",
    elev_p < 0.1 ~ "Marginal",
    TRUE ~ "Not significant"
  )

  model_summaries_S1D <- dplyr::bind_rows(
    model_summaries_S1D,
    data.frame(species = sp, p_value = elev_p, signif = signif_label)
  )

  mean_e <- mean(df$elev_indiv, na.rm = TRUE)
  sd_e <- sd(df$elev_indiv, na.rm = TRUE)

  elev_seq <- seq(
    min(df$elev_indiv, na.rm = TRUE),
    max(df$elev_indiv, na.rm = TRUE),
    length.out = 100
  )

  newdata <- data.frame(
    elev_indiv_scaled = (elev_seq - mean_e) / sd_e,
    diameter_complete_scaled = 0,
    canopy_closure_scaled = 0
  )

  newdata$elev_ft <- elev_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds_S1D <- dplyr::bind_rows(all_preds_S1D, newdata)
}

# ======================
# S1D: plot
# ======================

S1D <- ggplot() +
  geom_point(
    data = subset(trait_data1, species != "ABMA"),
    aes(x = elev_indiv, y = midday_MPa_2025, color = species),
    size = 3,
    alpha = 0.35
  ) +
  scale_color_manual(values = species_colors) +
  ggnewscale::new_scale_color() +
  geom_line(
    data = all_preds_S1D,
    aes(x = elev_ft, y = pred, color = species, linetype = signif),
    size = 1.2
  ) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant" = "solid",
    "Marginal" = "dashed",
    "Not significant" = "dotted"
  )) +
  labs(
    x = "Elevation (ft)",
    y = expression(Psi[midday]~"(MPa)"~"2025")
  ) + scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
) + 
  scale_x_continuous(
    breaks = seq(
      floor(min(trait_data1$elev_indiv, na.rm = TRUE)),
      ceiling(max(trait_data1$elev_indiv, na.rm = TRUE)),
      by = 500
    )
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    legend.position = "none"
  )

S1D


# ======================
# S1E: model + predictions
# ======================

all_preds_S1E <- data.frame()
model_summaries_S1E <- data.frame()

for (sp in setdiff(names(species_list), "ABMA")) {

  df <- species_list[[sp]]

  mod <- lmer(
    predawn_MPa_combined ~ elev_indiv_scaled + diameter_complete_scaled +
      canopy_closure_scaled + (1 | plot) + (1 | year),
    data = df
  )

  summ <- summary(mod)$coefficients
  elev_p <- summ["elev_indiv_scaled", "Pr(>|t|)"]

  signif_label <- dplyr::case_when(
    elev_p < 0.05 ~ "Significant",
    elev_p < 0.1 ~ "Marginal",
    TRUE ~ "Not significant"
  )

  model_summaries_S1E <- dplyr::bind_rows(
    model_summaries_S1E,
    data.frame(species = sp, p_value = elev_p, signif = signif_label)
  )

  mean_e <- mean(df$elev_indiv, na.rm = TRUE)
  sd_e <- sd(df$elev_indiv, na.rm = TRUE)

  elev_seq <- seq(
    min(df$elev_indiv, na.rm = TRUE),
    max(df$elev_indiv, na.rm = TRUE),
    length.out = 100
  )

  newdata <- data.frame(
    elev_indiv_scaled = (elev_seq - mean_e) / sd_e,
    diameter_complete_scaled = 0,
    canopy_closure_scaled = 0
  )

  newdata$elev_ft <- elev_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds_S1E <- dplyr::bind_rows(all_preds_S1E, newdata)
}

# ======================
# S1E: plot
# ======================

S1E <- ggplot() +
  geom_point(
    data = subset(trait_data1, species != "ABMA"),
    aes(x = elev_indiv, y = predawn_MPa_combined, color = species),
    size = 3,
    alpha = 0.35
  ) +
  scale_color_manual(values = species_colors) +
  ggnewscale::new_scale_color() +
  geom_line(
    data = all_preds_S1E,
    aes(x = elev_ft, y = pred, color = species, linetype = signif),
    size = 1.2
  ) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant" = "solid",
    "Marginal" = "dashed",
    "Not significant" = "dotted"
  )) +
  labs(
    x = "Elevation (ft)",
   y = expression(Psi[predawn]~"(MPa)"~"2-year average")
  ) + scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
) + 
  scale_x_continuous(
    breaks = seq(
      floor(min(trait_data1$elev_indiv, na.rm = TRUE)),
      ceiling(max(trait_data1$elev_indiv, na.rm = TRUE)),
      by = 500
    )
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    legend.position = "none"
  )

S1E


# ======================
# S1F: model + predictions
# ======================

all_preds_S1F <- data.frame()
model_summaries_S1F <- data.frame()

for (sp in setdiff(names(species_list), "ABMA")) {

  df <- species_list[[sp]]

  mod <- lmer(
    midday_MPa_combined ~ elev_indiv_scaled + diameter_complete_scaled +
      canopy_closure_scaled + (1 | plot) + (1 | year),
    data = df
  )

  summ <- summary(mod)$coefficients
  elev_p <- summ["elev_indiv_scaled", "Pr(>|t|)"]

  signif_label <- dplyr::case_when(
    elev_p < 0.05 ~ "Significant",
    elev_p < 0.1 ~ "Marginal",
    TRUE ~ "Not significant"
  )

  model_summaries_S1F <- dplyr::bind_rows(
    model_summaries_S1F,
    data.frame(species = sp, p_value = elev_p, signif = signif_label)
  )

  mean_e <- mean(df$elev_indiv, na.rm = TRUE)
  sd_e <- sd(df$elev_indiv, na.rm = TRUE)

  elev_seq <- seq(
    min(df$elev_indiv, na.rm = TRUE),
    max(df$elev_indiv, na.rm = TRUE),
    length.out = 100
  )

  newdata <- data.frame(
    elev_indiv_scaled = (elev_seq - mean_e) / sd_e,
    diameter_complete_scaled = 0,
    canopy_closure_scaled = 0
  )

  newdata$elev_ft <- elev_seq
  newdata$pred <- predict(mod, newdata, re.form = NA)
  newdata$species <- sp
  newdata$signif <- signif_label

  all_preds_S1F <- dplyr::bind_rows(all_preds_S1F, newdata)
}

# ======================
# S1F: plot
# ======================

S1F <- ggplot() +
  geom_point(
    data = subset(trait_data1, species != "ABMA"),
    aes(x = elev_indiv, y = midday_MPa_combined, color = species),
    size = 3,
    alpha = 0.35
  ) +
  scale_color_manual(values = species_colors) +
  ggnewscale::new_scale_color() +
  geom_line(
    data = all_preds_S1F,
    aes(x = elev_ft, y = pred, color = species, linetype = signif),
    size = 1.2
  ) +
  scale_color_manual(values = species_colors) +
  scale_linetype_manual(values = c(
    "Significant" = "solid",
    "Marginal" = "dashed",
    "Not significant" = "dotted"
  )) +
  labs(
    x = "Elevation (ft)",
    y = expression(Psi[midday]~"(MPa)"~"2-year average")
  ) +
  theme_bw() + scale_y_reverse(
  labels = function(x) paste0("\u2212", x)
) + 
  scale_x_continuous(
    breaks = seq(
      floor(min(trait_data1$elev_indiv, na.rm = TRUE)),
      ceiling(max(trait_data1$elev_indiv, na.rm = TRUE)),
      by = 500
    )
  ) +
  theme(
    panel.grid = element_blank(),
    legend.position = "none"
  )

S1F

grid.arrange(S1A, S1B, S1C, S1D, S1E, S1F, nrow = 3, ncol = 2)


```



# stats - Table 1, 2, 3
```{r}

mod <- lmer(internode_length_1_2~ diameter_complete_scaled + predawn_scaled + HSM_midday_P50_mean_scaled + canopy_closure_scaled + elev_indiv_scaled + (1|plot), data = PILA)
summary(mod)


mod1 <- lm(HSM_predawn_combined ~ diameter, data = ABCO)
summary(mod1)

t.test(predawn_MPa_combined ~ year, data = trait_data1, paired = FALSE)
# There is a statistically significant difference between 2024 and 2025 predawn and midday water potential means. On average, predawn WP is higher (more negative) in 2024 and midday is 


trait_data1$HSM_predawn_scaled <- as.numeric(scale(trait_data1$HSM_predawn))
trait_data1$HSM_predawn_P50_mean_scaled <- as.numeric(scale(trait_data1$HSM_predawn_P50_mean))
trait_data1$HSM_midday_P50_mean_scaled <- as.numeric(scale(trait_data1$HSM_midday_P50_mean))

# run models with scaled response and predictor variables
mod5 <- lmer(HSM_predawn_P50_mean ~DBH_complete_scaled * percent_closure_scaled * elev_indiv_scaled + (1|plot), data = PIPO)
summary(mod5)

install.packages("emmeans")
library(emmeans)

mod4 <- lmer(HSM_midday_P50_mean ~ species * diameter_complete_scaled * elev_indiv_scaled * canopy_closure_scaled + (1|plot) + (1|year), data = trait_data1)
summary(mod4)

# table 3 pairwise comparisons using emmeans
emms1 <- emmeans(mod4, pairwise ~ species)
summary(emms1$emmeans)
summary(emms1$contrasts) #automatically adjusts to tukey, for multiple comparisons



model <- lmer(internode_length_1_2 ~  predawn_scaled + canopy_closure_scaled + P50_scaled +(1|plot), data = PILA)
summary(model)

trait_data1$residuals <- resid(model)





```


# AICc model selection
```{r}
install.packages('MuMIn')
library(MuMIn)

dredge_df <- na.omit(ABCO[, c("HSM_midday", "elev_indiv", "DBH_complete", "percent_cover", "plot")])
global <- lmer(HSM_midday ~ elev_indiv * DBH_complete * percent_cover + (1|plot), data = dredge_df, na.action = na.fail)
dredge(global)

# likelihood ratio test for comparison of models by hypothesis testing versus AICc model selection
mod6<- lmer(HSM_predawn_P50_mean_scaled ~ 1 + (1|plot), data = dredge_df)
mod7 <- lmer(HSM_predawn_P50_mean_scaled ~elev_indiv_scaled + (1|plot), data = dredge_df)
mod8 <- lmer(HSM_predawn_P50_mean_scaled ~DBH_complete_scaled + elev_indiv_scaled + (1|plot), data = dredge_df)
mod9 <- lmer(HSM_predawn_P50_mean_scaled ~DBH_complete_scaled + elev_indiv_scaled + percent_closure_scaled +  (1|plot), data = dredge_df)

anova(mod6, mod7, mod8, mod9)
```

