---
title: "07_preliminary-analyses"
author: "Anjum K. Gujral"
date: '2024-11-11'
output:
  pdf_document:
    output_dir: "/Users/anjumgujral/Documents/repos/yuba-phys"
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
trait_data1 <- read.csv("/Users/anjumgujral/Documents/repos/yuba-phys/trait_data1.csv")

install.packages("tinytex")
library('tinytex')
library('dplyr')
library('tidyverse')
library('ggplot2')

```

Size distributions of seedlings and saplings

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = DBH_complete, fill = size_class)) +
  geom_histogram(binwidth = 0.5, position = "dodge", alpha = 0.7, color = "black") +
  labs( 
       x = "Diameter at root collar (cm)", 
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "lightgreen")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

Distribution of seedlings and saplings across light environments
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = DBH_complete, y = percent_cover, color = size_class)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
       y = "Canopy closure", 
       x = "Diameter at root collar") +
  scale_color_manual(values = c("blue", "green")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

Distribution of microsite environments across elevation
-- how similar are microsites at high and low elevation 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = percent_cover, fill = microsite)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Canopy Closure"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))
```

Distribution of stand basal area across elevation
-- how similar is stand density between plots at high and low elevation
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = stand_basal_area, fill = elevation)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Stand Basal Area"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))
```

Do seedlings and saplings differ in predawn and midday water potentials across elevation?
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = size_class, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())
```

```{r}
### stats
mod1 <- lm(predawn_MPa ~ species * size_class * elevation, 
             data = trait_data1)
summary(mod1)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = size_class, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod2 <- lm(midday_MPa ~ species * size_class * elevation, 
             data = trait_data1)
summary(mod2)

```

Do predawn and midday water potentials differ across microsite?
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod3 <- lm(predawn_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod3)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod4 <- lm(midday_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod4)

```

Do predawn and midday water potentials differ by microsite, species, and elevation?
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = microsite, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
       x = "Microsite",
       y = "Predawn Water Potential (-MPa)",
       fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
mod5 <- lm(predawn_MPa ~ species * microsite * elevation, 
             data = trait_data1)
summary(mod5)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
       x = "Microsite",
       y = "Midday Water Potential (-MPa)",
       fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
mod6 <- lm(midday_MPa ~ species * microsite * elevation, 
             data = trait_data1)
summary(mod6)
```
