---
title: "07_preliminary-analyses"
author: "Anjum K. Gujral"
date: '2024-11-11'
output:
  html_document:
    df_print: paged
  pdf_document:
    output_dir: /Users/anjumgujral/Documents/repos/yuba-phys
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
trait_data1 <- read.csv("/Users/anjumgujral/Documents/repos/yuba-phys/trait_data1.csv")

install.packages("tinytex")
library('tinytex')
library('dplyr')
library('tidyverse')
library('ggplot2')

```

Size distributions of seedlings and saplings

{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = DBH_complete, fill = size_class)) +
  geom_histogram(binwidth = 0.5, position = "dodge", alpha = 0.7, color = "black") +
  labs( 
       x = "Diameter at root collar (cm)", 
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "lightgreen")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())


Distribution of seedlings and saplings across light environments
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = DBH_complete, y = percent_cover, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
       y = "Canopy closure", 
       x = "Diameter at root collar") +
  scale_color_manual(values = c("blue", "green", "red", "purple")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

# Update rmarkdown
install.packages("rmarkdown")

# Update xfun
install.packages("xfun")

```

Distribution of microsite environments across elevation
-- how similar are microsites at high and low elevation 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = percent_cover, fill = microsite)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Canopy Closure"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))


# Create a new variable combining size_class and elevation
trait_data1$size_elevation <- interaction(trait_data1$size_class, trait_data1$elevation)

# Create the plot
ggplot(trait_data1, aes(x = species, y = percent_cover)) +
  # Create one boxplot for canopy closure by species
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  
  # Add points, color-coded by the combination of size and elevation
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  
  # Axis labels
  labs(
    y = "Canopy closure", 
    x = "Species",
    color = "Size & Elevation"  # Label for the combined legend
  ) +
  
  # Clean theme
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )



# Create the plot
ggplot(trait_data1, aes(x = microsite, y = percent_cover)) +
  # Create one boxplot for each microsite level
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  
  # Add points, color-coded by the combination of size and elevation
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  
  # Axis labels
  labs(
    y = "Canopy closure", 
    x = "Microsite level", 
    color = "Size & Elevation"  # Label for the combined legend
  ) +
  # Facet by species: create a separate plot for each species
  facet_wrap(~ species, scales = "free_y") +
  
  # Clean theme
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```

Distribution of stand basal area across elevation
-- how similar is stand density between plots at high and low elevation
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = stand_basal_area, fill = elevation)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Stand Basal Area"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))
```

Do seedlings and saplings differ in predawn and midday water potentials across elevation?

```{r}
ggplot(trait_data1, aes(x = DBH_complete, y = predawn_MPa, color = species)) +
  geom_point(size = 3) +                # Adds scatter plot points
  geom_smooth(method = "lm", se = FALSE, aes(group = species), linetype = "solid") +  # Add regression lines
  labs(
    title = "Predawn Water Potential by Tree Diameter",
    x = "Diameter at root collar (cm)",
    y = "Predawn Water Potential (-MPa)",
    color = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "top")

```


```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = size_class, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())
```

```{r}
### stats
mod1 <- lm(predawn_MPa ~ species * size_class * elevation, 
             data = trait_data1)
summary(mod1)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = size_class, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod2 <- lm(midday_MPa ~ species * size_class * elevation, 
             data = trait_data1)
summary(mod2)

```

Do predawn and midday water potentials differ across microsite?
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod3 <- lm(predawn_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod3)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod4 <- lm(midday_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod4)

```

Do predawn and midday water potentials differ across elevation?
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Elevation",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod3 <- lm(predawn_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod3)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod4 <- lm(midday_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod4)

```

Do predawn and midday water potentials differ by microsite, species, and elevation?
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = microsite, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
       x = "Microsite",
       y = "Predawn Water Potential (-MPa)",
       fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
mod5 <- lm(predawn_MPa ~ species * microsite * elevation, 
             data = trait_data1)
summary(mod5)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
       x = "Microsite",
       y = "Midday Water Potential (-MPa)",
       fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
mod6 <- lm(midday_MPa ~ species * microsite * elevation, 
             data = trait_data1)
summary(mod6)
```

```{r}
library(lme4)
library(lmerTest)

ABCO <- trait_data1[trait_data1$species == 'ABCO',]
PILA <- trait_data1[trait_data1$species == 'PILA',]
PIPO <- trait_data1[trait_data1$species == 'PIPO',]
ABMA <- trait_data1[trait_data1$species == 'ABMA',]

mod7<- lmer(ABCO$midday_MPa~ microsite + size_class + microsite:size_class + (1|plot), REML = TRUE, data = ABCO)
summary(mod7)

```

# turning linear models into fitted models 
# run your lm() or gml() model
# The fitted values (predictions) are the values predicted by the model for each observation in the dataset based on the estimated model parameters. You can access them using the fitted() function or the $fitted.values attribute of the model object.





