---
title: "07_preliminary-analyses"
author: "Anjum K. Gujral"
date: '2024-11-11'
output:
  html_document:
    df_print: paged
  pdf_document:
    output_dir: /Users/anjumgujral/Documents/repos/yuba-phys
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
trait_data1 <- read.csv("/Users/anjumgujral/Documents/repos/yuba-phys/trait_data1.csv")

install.packages("tinytex")
library('tinytex')
library('dplyr')
library('tidyverse')
library('ggplot2')
library(emmeans)
library(lme4)

```

#size distribution
```{r, echo = FALSE, message = FALSE, warning = FALSE}
# create a new column for size that categorizes individuals with diameter at root collar > 2.5 cm as saplings and individuals with diameter at root collar < 2.5 cm seedlings



ggplot(trait_data1, aes(x = DBH_complete, fill = size)) +
  geom_histogram(binwidth = 0.5, position = "dodge", alpha = 0.7, color = "black") +
  labs( 
       x = "Diameter at root collar (cm)", 
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "lightgreen")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

# microsite distributions
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = DBH_complete, y = percent_cover, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
       y = "Canopy closure", 
       x = "Diameter at root collar") +
  scale_color_manual(values = c("blue", "green", "red", "purple")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

Distribution of microsite environments across elevation
-- how similar are microsites at high and low elevation 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = percent_cover, fill = microsite)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Canopy Closure"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))

# Create a new variable combining size_class and elevation
trait_data1$size_elevation <- interaction(trait_data1$size, trait_data1$elevation)


ggplot(trait_data1, aes(x = species, y = percent_cover)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Canopy closure", 
    x = "Species",
    color = "Size & Elevation"  
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )


ggplot(trait_data1, aes(x = microsite, y = percent_cover)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Canopy closure", 
    x = "Microsite level", 
    color = "Size & Elevation"  
  ) +
  facet_wrap(~ species, scales = "free_y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```

# stand basal area
Distribution of stand basal area across elevation
-- how similar is stand density between plots at high and low elevation
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = stand_basal_area, fill = elevation)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Stand Basal Area"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))



ggplot(trait_data1, aes(x = species, y = stand_basal_area)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Stand Basal Area ", 
    x = "Species",
    color = "Size & Elevation"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )


ggplot(trait_data1, aes(x = microsite, y = stand_basal_area)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Stand Basal Area", 
    x = "Microsite level", 
    color = "Size & Elevation" 
  ) +
  facet_wrap(~ species, scales = "free_y") +
  
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```

# environmental variables
```{r}
# Create a correlation matrix for environmental variables
cor_matrix <- cor(trait_data1[, c("percent_cover", "elev_band_ft", "stand_basal_area")], use = "pairwise.complete.obs")
print(cor_matrix)

cor(trait_data1$elev_band_ft, trait_data1$stand_basal_area, method = "pearson")
cor.test(trait_data1$stand_basal_area, trait_data1$percent_cover, method="pearson")

install.packages("ggcorrplot") 
library(ggcorrplot)
library(corrplot)
library("ggpubr")

# Create a custom set of axis labels (modify as per your needs)
custom_labels <- c("Canopy Closure", "Elevation (ft)", "Stand Basal Area")

colnames(cor_matrix) <- custom_labels
rownames(cor_matrix) <- custom_labels

# Visualize the correlation matrix with custom axis labels
corrplot(cor_matrix, 
         method = "color",          
         type = "lower",            
         order = "hclust",          
         addCoef.col = "black",    
         tl.col = "black",          
         tl.srt = 45,               
         tl.labels = custom_labels, 
         title = "Correlation Plot"
)

```

# predawn
```{r}

p1 <- ggplot(subset(trait_data1, species != "ABMA"), aes(x = elev_indiv, y = HSM_predawn_P50_mean, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "HSM (MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

p1

p2 <- ggplot(subset(trait_data1, species != "ABMA" & !is.na(size)), aes(x = elev_indiv, y = HSM_predawn_P50_mean, color = species)) + facet_wrap(~size) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "HSM (MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse() + 
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                by = 500)) +
theme(
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  legend.position = "none"
)

p2
install.packages('MuMIn')
library(MuMIn)
dredge_df <- na.omit(PIPO[, c("predawn_MPa", "elev_indiv", "DBH_complete", "percent_cover", "plot")])
global <- lmer(predawn_MPa ~ elev_indiv * DBH_complete * percent_cover + (1|plot), data = dredge_df, na.action = na.fail)
dredge(global)

mod1 <- lm(predawn_MPa~DBH_complete + elevation, data = trait_data1)
colors <- ifelse (trait_data1$elevation=='low', "black", "gray")
plot (trait_data1$DBH_complete, trait_data1$predawn_MPa, xlab="Diameter at root collar", ylab="Predawn water potential",
col=colors, pch=20)
curve (cbind (1, 1, x) %*% coef(mod1), add=TRUE, col="black")
curve (cbind (1, 0, x) %*% coef(mod1), add=TRUE, col="gray")

plot(trait_data1$DBH_complete, trait_data1$predawn_MPa, xlab="Diameter at root collar", ylab="Predawn water potential")
curve (coef(mod1)[1] + coef(mod1)[2]*x, add=TRUE) ## plotting the fitted regression line from the linear model rather than ggplot


trait_data2 <- trait_data1 %>%
  mutate(time_of_day = "predawn", water_potential = predawn_MPa) %>%
  bind_rows(
    trait_data1 %>%
      mutate(time_of_day = "midday", water_potential = midday_MPa)
  )


install.packages("gridExtra")
library(gridExtra)
grid.arrange(p1, p2, nrow = 2, ncol = 1)
```


Do seedlings and saplings differ in predawn and midday water potentials across elevation?

```{r}
ggplot(trait_data1, aes(x = DBH_complete, y = predawn_MPa, color = species)) +
  geom_point(size = 3) +                
  geom_smooth(method = "lm", se = FALSE, aes(group = species), linetype = "solid") +  
  labs(
    title = "Predawn Water Potential by Tree Diameter",
    x = "Diameter at root collar (cm)",
    y = "Predawn Water Potential (-MPa)",
    color = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "top")

```


```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = size, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())
```

```{r}
### stats
mod1 <- lmer(predawn_MPa ~ DBH_complete + elevation + size:elevation +(1|plot), 
             data = trait_data2)
summary(mod1)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = size, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_fill_manual(values = c("ABCO" = "blue", "PIPO" = "green", "ABMA" = "red", "PILA" = "purple")) + 
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

## predawn by species, size and elevation
ggplot(subset(trait_data1, !is.na(size)), aes(x = size, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  geom_jitter(color = "black",
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
              size = 1.5, alpha = 0.6, show.legend = FALSE) +
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_fill_manual(values = c("ABCO" = "blue", "PIPO" = "green", "ABMA" = "red", "PILA" = "purple")) + 
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())


mod7<- lmer(ABCO$predawn_MPa ~ DBH_complete + percent_cover + (1|plot), REML = TRUE, data = ABCO)
summary(mod7)

# predawn by species, size, and microsite
ggplot(within(subset(trait_data1, !is.na(microsite) & !is.na(size)), 
              microsite <- factor(microsite, levels = c("sun", "int", "shade"))),
       aes(x = size, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  geom_jitter(color = "black",
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
              size = 1.5, alpha = 0.6, show.legend = FALSE) +
  facet_wrap(~ microsite) +
  labs(
    x = "Size",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_fill_manual(values = c("ABCO" = "blue", "PIPO" = "green", "ABMA" = "red", "PILA" = "purple")) + 
  scale_y_reverse() +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())



ggplot(subset(trait_data1, !is.na(size)), aes(x = size, y = DBH_complete)) +
  geom_boxplot(fill = "white", color = "black", outlier.shape = NA) +
  geom_jitter(aes(color = species),
              width = 0.2, alpha = 0.6, size = 1.8, show.legend = TRUE) +
  labs(
    x = "Size Class",
    y = "Diameter at root collar (cm)",
    color = "Species"
  ) +
  scale_color_manual(values = c("ABCO" = "blue", "PIPO" = "green", "ABMA" = "red", "PILA" = "purple")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )





library(dplyr)

# Remove rows with NA values in 'microsite'
trait_data1_microsite <- trait_data1 %>%
  filter(!is.na(microsite))


```

```{r}
### stats
mod2 <- lm(midday_MPa ~ species * size_class * elevation, 
             data = trait_data1)
summary(mod2)

```

Do predawn and midday water potentials differ across microsite?
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = size, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod3 <- lm(predawn_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod3)

```

# midday
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod4 <- lm(midday_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod4)

```

Do predawn and midday water potentials differ across elevation?
```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
library(viridis)
ggplot(trait_data1, aes(x = elevation, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Elevation",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  scale_fill_viridis(discrete = TRUE) + 
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod3 <- lm(predawn_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod3)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod4 <- lm(midday_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod4)

```

Do predawn and midday water potentials differ by microsite, species, and elevation?
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = microsite, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
       x = "Microsite",
       y = "Predawn Water Potential (-MPa)",
       fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

# P50
```{r}
library(ggplot2)

ggplot(subset(trait_data2, !is.na(microsite) & !is.na(size)), aes(x = microsite, y = HSM_predawn_P50_mean, fill = size)) + 
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +
  geom_jitter(position = position_jitter(width = 0.2), color = "black", alpha = 0.6, size = 1.5) +
  #scale_fill_manual(values = c(
    #"seedling" = "skyblue",
    #"sapling" = "lightgreen"
  #)) +
  labs(
    x = "Microsite",
    y = "HSM (MPa)",
    fill = "Size"
  ) +
  facet_wrap(~ species) +  
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),  
    strip.text = element_text(size = 12)
  )
mod1<- lm(P50_MPa ~ DBH_complete, data = PIPO)
summary(mod1)

### stats
mod5 <- lm(predawn_MPa ~  elev_indiv + percent_cover + DBH_complete, data = ABCO)
summary(mod5)

ggplot(trait_data2[!is.na(trait_data2$size), ], aes(y = P50_MPa, x = percent_cover, color = species)) +
  geom_point(size = 3, alpha = 1) +
  facet_wrap(~size)+
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +  
  labs(
    x = "Percent closure",
    y = "P50 (-MPa)",
    color = "Species"
  ) +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black")
  )

ggplot(subset(trait_data1,!is.na(size)), aes(x = percent_cover, y = HSM_predawn_P50_mean, color = species)) +
  facet_wrap(~size) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  #scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                  #ceiling(max(trait_data1$elev_indiv)),
                                 # by = 500)) +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"  
  )) +
  labs(x = "Percent closure", y = "HSM (MPa)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black")
  )


p2 <- ggplot(trait_data1, aes(x = DBH_complete, y = HSM_predawn_P50_mean, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  #scale_x_continuous(breaks = #seq(floor(min(trait_data1$elev_indiv)),                              #ceiling(max(trait_data1$elev_indiv)),
                                  #by = 500)) +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  labs(x = "Diameter at root collar (cm)", y = "HSM (MPa)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    legend.position = "none",
    axis.line = element_line(color = "black")
  )

p1 
grid.arrange(p1, p2, nrow = 2, ncol = 1)

mod6<- lm(trait_data2$midday_MPa~DBH_complete, data= trait_data2)
summary(mod6)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
       x = "Microsite",
       y = "Midday Water Potential (-MPa)",
       fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
mod6 <- lm(elevation ~ species + microsite + elevation, 
             data = trait_data1)
summary(mod6)
```

```{r}


```


# HSM
```{r}

library(tidyverse)

trait_data2 <- trait_data1 %>%
  pivot_longer(
    cols = c(predawn_MPa, midday_MPa, P50_MPa, P50_mean, HSM_predawn, HSM_midday),
    names_to = "type",
    values_to = "water_potential"
  )

ggplot(data = subset(trait_data2, !type %in% c("HSM_predawn", "HSM_midday")),
       aes(x = elev_indiv, y = water_potential, color = type)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c(
    "P50_MPa" = "darkslategray4",
    "midday_MPa" = "cadetblue3",
    "predawn_MPa" = "steelblue3"
  )) +
  theme_minimal() +
  labs(
    x = "Elevation",
    y = "Water Potential (-MPa)",
    color = "Type"
  ) + 
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )





ggplot(data = subset(trait_data2, !type %in% c("HSM_predawn", "HSM_midday",  "P50_MPa")),
       aes(x = elev_indiv, y = water_potential, color = type)) +
  facet_wrap(~species) +
  geom_line(data = subset(trait_data2, type %in% c("predawn_MPa", "P50_mean")),
            aes(x = elev_indiv, y = water_potential, group = tree_ID),
            color = "gray40",
            size = 0.6,
            inherit.aes = FALSE) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)), ceiling(max(trait_data1$elev_indiv)), by = 500)) +
  scale_color_manual(values = c(
    "P50_mean" = "darkblue",
    "midday_MPa" = "darkcyan",
    "predawn_MPa" = "steelblue3"
  )) +
  theme_bw() +
  labs(
    title = "P50 means",
    x = "Elevation",
    y = "Water Potential (-MPa)",
    color = "type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )




ggplot(data = subset(trait_data2, !type %in% c("HSM_predawn", "HSM_midday")),
       aes(x = DBH_complete, y = water_potential, color = type)) +
  facet_wrap(~species) +
  geom_line(data = subset(trait_data2, type %in% c("predawn_MPa", "P50_MPa")),
            aes(x = DBH_complete, y = water_potential, group = tree_ID),
            color = "gray40",
            size = 0.6,
            inherit.aes = FALSE) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c(
    "P50_mean" = "darkblue",
    "midday_MPa" = "darkcyan",
    "predawn_MPa" = "steelblue3"
  )) +
  theme_bw() +
  labs(
    title = "using actual data",
    x = "Diameter at root collar (cm)",
    y = "Water Potential (-MPa)",
    color = "type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

```


```{r}
library(lme4)
library(lmerTest)

ABCO <- trait_data1[trait_data1$species == 'ABCO',]
PILA <- trait_data1[trait_data1$species == 'PILA',]
PIPO <- trait_data1[trait_data1$species == 'PIPO',]
ABMA <- trait_data1[trait_data1$species == 'ABMA',]


trait_data1$stand_basal_area
ABCO_cleaned_data <- ABCO[!is.na(ABCO$DBH_complete) & !is.na(ABCO$percent_cover) & !is.na(ABCO$y), ]
ABCO_cleaned_data <- ABCO[!is.na(ABCO$DBH_complete) & !is.na(ABCO$percent_cover), ]


mod7<- lmer(ABCO$midday_MPa~ DBH_complete + percent_cover +DBH_complete:percent_cover + (1|plot), REML = TRUE, data = PILA)
summary(mod7)

#Get the fitted values (predicted values) from the model
fitted_values <- fitted(mod7)

# Create a new dataframe with predictors and fitted values
fitted_data <- data.frame(DBH = ABCO_cleaned_data$DBH_complete, 
                        percent_closure = ABCO_cleaned_data$percent_cover, 
                        fitted_values = fitted_values)

library(ggplot2)
ggplot(plot_data, aes(x = x1, y = fitted_values, color = x2)) +
  geom_line() +  # Fit line
  geom_point(data = your_data, aes(x = x1, y = y), color = "red", alpha = 0.5) +  # Raw data as points
  labs(title = "Fitted Values from Linear Mixed Model",
       x = "Predictor 1 (x1)", y = "Fitted Values (y)") +
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient for x2
  theme_minimal()

# Create a new dataframe that excludes species ABMA to look at elevation
trait_data2 <- trait_data1[trait_data1$species != "ABMA", ]

mod8<- lm(trait_data2$elev_band_ft ~ stand_basal_area + species + percent_cover + stand_basal_area:species + percent_cover:species, data = trait_data2)
summary(mod8) 

# Compute estimated marginal means for percent_cover
emm_percent_cover <- emmeans(mod7, ~ percent_cover)

# Compute estimated marginal means for DBH_complete (with percent_cover held constant)
emm_DBH <- emmeans(mod7, ~ DBH_complete)

# Alternatively, for interaction, you can compute both percent_cover and DBH_complete effects together
emm_interaction <- emmeans(mod7, ~ microsite * size)
summary(emm_interaction)
emm_result_df <- as.data.frame(emm_interaction)

# Plot the effect of percent_cover on midday_MPa
ggplot(emm_percent_cover, aes(x = percent_cover, y = emmean, ymin = lower.CL, ymax = upper.CL)) +
  geom_point() +
  geom_errorbar(width = 0.2) +  # Add error bars for confidence intervals
  labs(x = "Percent Cover", y = "Estimated Marginal Mean of Midday MPa") +
  theme_minimal() +
  ggtitle("Effect of Percent Cover on Midday MPa")

# Plot the interaction between percent_cover and DBH_complete
ggplot(emm_result_df, aes(x = size, y = emmean, color = microsite)) +
  geom_line() + 
  geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL, fill = microsite), alpha = 0.3) + 
  labs(x = "size", y = "Estimated Marginal Mean of Midday MPa") +
  #scale_color_viridis_c() +  # Color scale for percent_cover
  #scale_fill_viridis_c() +  # Color scale for the ribbon fill
  theme_minimal() +
  ggtitle("Interaction effect between microsite and size on midday WP")

# Inspect the structure of your emm_result_df
str(emm_result_df)

# Check the first few rows of the dataframe
head(emm_result_df)

# Summary statistics to see if values are within expected ranges
summary(emm_result_df)

```





