---
title: "07_preliminary-analyses"
author: "Anjum K. Gujral"
date: '2024-11-11'
output:
  html_document:
    df_print: paged
  pdf_document:
    output_dir: /Users/anjumgujral/Documents/repos/yuba-phys
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
trait_data1 <- read.csv("/Users/anjumgujral/Documents/repos/yuba-phys/trait_data1.csv")

install.packages("tinytex")
library('tinytex')
library('dplyr')
library('tidyverse')
library('ggplot2')
library(emmeans)
library(lme4)

```

#size distribution
```{r, echo = FALSE, message = FALSE, warning = FALSE}
# create a new column for size that categorizes individuals with diameter at root collar > 2.5 cm as saplings and individuals with diameter at root collar < 2.5 cm seedlings



ggplot(trait_data1, aes(x = DBH_complete, fill = size)) +
  geom_histogram(binwidth = 0.5, position = "dodge", alpha = 0.7, color = "black") +
  labs( 
       x = "Diameter at root collar (cm)", 
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "lightgreen")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

# microsite distributions
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = DBH_complete, y = percent_cover, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
       y = "Canopy closure", 
       x = "Diameter at root collar") +
  scale_color_manual(values = c("blue", "green", "red", "purple")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

Distribution of microsite environments across elevation
-- how similar are microsites at high and low elevation 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = percent_cover, fill = microsite)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Canopy Closure"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))

# Create a new variable combining size_class and elevation
trait_data1$size_elevation <- interaction(trait_data1$size, trait_data1$elevation)


ggplot(trait_data1, aes(x = species, y = percent_cover)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Canopy closure", 
    x = "Species",
    color = "Size & Elevation"  
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )


ggplot(trait_data1, aes(x = microsite, y = percent_cover)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Canopy closure", 
    x = "Microsite level", 
    color = "Size & Elevation"  
  ) +
  facet_wrap(~ species, scales = "free_y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```

# stand basal area
Distribution of stand basal area across elevation
-- how similar is stand density between plots at high and low elevation
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = stand_basal_area, fill = elevation)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Stand Basal Area"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))



ggplot(trait_data1, aes(x = species, y = stand_basal_area)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Stand Basal Area ", 
    x = "Species",
    color = "Size & Elevation"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )


ggplot(trait_data1, aes(x = microsite, y = stand_basal_area)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Stand Basal Area", 
    x = "Microsite level", 
    color = "Size & Elevation" 
  ) +
  facet_wrap(~ species, scales = "free_y") +
  
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```

# environmental variables
```{r}
# Create a correlation matrix for environmental variables
cor_matrix <- cor(trait_data1[, c("percent_cover", "elev_band_ft", "stand_basal_area")], use = "pairwise.complete.obs")
print(cor_matrix)

cor(trait_data1$elev_band_ft, trait_data1$stand_basal_area, method = "pearson")
cor.test(trait_data1$stand_basal_area, trait_data1$percent_cover, method="pearson")

install.packages("ggcorrplot") 
library(ggcorrplot)
library(corrplot)
library("ggpubr")

# Create a custom set of axis labels (modify as per your needs)
custom_labels <- c("Canopy Closure", "Elevation (ft)", "Stand Basal Area")

colnames(cor_matrix) <- custom_labels
rownames(cor_matrix) <- custom_labels

# Visualize the correlation matrix with custom axis labels
corrplot(cor_matrix, 
         method = "color",          
         type = "lower",            
         order = "hclust",          
         addCoef.col = "black",    
         tl.col = "black",          
         tl.srt = 45,               
         tl.labels = custom_labels, 
         title = "Correlation Plot"
)



```

# Figure 1 - size
```{r}

s1 <- ggplot(trait_data1, aes(y = predawn_MPa_combined, x = diameter_complete, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +  
  labs(
    x = "Diameter at root collar",
    y = "Predawn water potential (-MPa)",
    color = "Species"
  ) +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
  legend.position = "none"
  )

s1

s2 <- ggplot(trait_data1, aes(x = diameter_complete, y = midday_MPa_combined, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  labs(x = "Diameter at root collar (cm)", y = "Midday water potential (-MPa)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
  legend.position = "none"
  )

s2 

s3 <- ggplot(trait_data2, aes(x = diameter_complete, y = P50_complete, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  labs(x = "Diameter at root collar (cm)", y = "P50 (-MPa)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "none"
  )

s3

s6 <- ggplot() +
  # 2024 points and smooth
  geom_point(data = trait_data1, 
             aes(x = diameter_complete, y = P50_MPa_2024, color = species), 
             size = 3, alpha = 1, shape = 16) +
  geom_smooth(data = trait_data1, 
              aes(x = diameter_complete, y = P50_MPa_2024, color = species), 
              method = "lm", se = TRUE, alpha = 0.3) +
  
  # 2025 points and smooth
  geom_point(data = trait_data1, 
             aes(x = diameter_complete, y = P50_MPa_2025, color = species), 
             size = 3, alpha = 1, shape = 17) +
  geom_smooth(data = trait_data1, 
              aes(x = diameter_complete, y = P50_MPa_2025, color = species), 
              method = "lm", se = TRUE, alpha = 0.3, linetype = "dashed") +

  # Custom colors
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +

  # Labels and theme
  labs(x = "Diameter at root collar (cm)", y = "P50 (-MPa)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "none"
  )

s6



s4 <- ggplot(trait_data2, aes(x = diameter_complete, y = HSM_predawn_P50_mean, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  labs(x = "Diameter at root collar (cm)", y = "HSM predawn (MPa)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "none"
  )

s4

s5 <- ggplot(trait_data2, aes(x = diameter_complete, y = HSM_midday_P50_mean, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  labs(x = "Diameter at root collar (cm)", y = "HSM midday (MPa)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "none"
  )

s5
grid.arrange(s1, s2, s4, s5, s3, nrow = 3, ncol = 2)

```
# Figure 2- predawn

```{r}

r1 <- ggplot(trait_data2, aes(y = HSM_midday_P50_mean, x = predawn_MPa_combined, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +  
  labs(
    x = "Predawn Water Potential (-MPa)",
    y = "HSM Midday (MPa)",
    color = "Species"
  ) +
    theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

r1


r2 <- ggplot(trait_data2, aes(y = P50_complete, x = predawn_MPa_combined, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +  
  labs(
    x = "Predawn Water Potential (-MPa)",
    y = "P50 (-MPa)",
    color = "Species"
  ) +
    theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

r2


grid.arrange(r1, r2, nrow = 2, ncol = 1)


plot(trait_data1$elev_indiv, trait_data1$midday_MPa_2024,
     col = "blue", pch = 19,
     xlab = "Elevation (ft)", ylab = "Midday Water Potential (-MPa)",
     xlim = range(trait_data1$elev_indiv, na.rm = TRUE),
     ylim = range(c(trait_data1$midday_MPa_2024, trait_data1$midday_MPa_2025), na.rm = TRUE))

# Add 2025 points
points(trait_data1$elev_indiv, trait_data1$midday_MPa_2025, col = "plum3", pch = 19)

# Add linear trendline for 2024
mod_2024_pd <- lm(midday_MPa_2024 ~ elev_indiv, data = trait_data1)
abline(mod_2024_pd, col = "blue", lwd = 2)

# Add linear trendline for 2025
mod_2025_pd <- lm(midday_MPa_2025 ~ elev_indiv, data = trait_data1)
abline(mod_2025_pd, col = "plum3", lwd = 2)


legend("topleft", legend = c("2024", "2025"),
       col = c("blue", "plum3"), pch = 19, lty = 1, lwd = 2, cex = 1)



plot(trait_data1$elevation, trait_data1$predawn_MPa_2024, type = "o", col = "blue",
     ylim = range(c(trait_data1$predawn_MPa_2024, trait_data1$predawn_MPa_2025), na.rm = TRUE),
     xlab = "Elevation", ylab = "Predawn Water Potential",
     main = "Predawn Water Potential vs Elevation")

# Add predawn_2025 to same plot
lines(df$elevation, df$predawn_2025, type = "o", col = "red")
legend("bottomleft", legend = c("2024", "2025"), col = c("blue", "red"), lty = 1, pch = 1)

# Midday plot
plot(df$midday_2024, type = "o", col = "blue",
     ylim = range(c(df$midday_2024, df$midday_2025), na.rm = TRUE),
     ylab = "Water Potential", xlab = "Sample Index", main = "Midday Water Potential")
lines(df$midday_2025, type = "o", col = "red")
legend("topright", legend = c("2024", "2025"), col = c("blue", "red"), lty = 1, pch = 1)



```

# Figure 3 - elev & microsite
```{r}

m1 <- ggplot(trait_data2, aes(x = percent_cover, y = P50_complete, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Canopy closure (%)", 
    y = "P50 (-MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse() + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

m1

m2 <- ggplot(trait_data2, aes(x = percent_cover, y = HSM_midday_P50_mean, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Canopy closure (%)", 
    y = "HSM Midday (MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  #scale_y_reverse() + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

m2

e1 <- ggplot(subset(trait_data2, species != "ABMA"), aes(x = elev_indiv, y = HSM_midday_P50_mean, color = species))+
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "HSM midday", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

e1

e2 <- ggplot(subset(trait_data2, species != "ABMA"), aes(x = elev_indiv, y = P50_MPa, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE,  alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "P50 (MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)),
                                ceiling(max(trait_data1$elev_indiv)),
                                 by = 500)) +
  #scale_y_reverse() + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

e2


grid.arrange(r1, r2, m2, m1, e2, e1, nrow = 3, ncol = 2)

e3 <- ggplot(
  subset(trait_data1, species != "ABMA"),
  aes(x = elev_indiv, y = midday_MPa_2025, color = species)
) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.25) +
  labs(
    x = "Elevation (ft)", 
    y = "Midday WP (-MPa)", 
    color = "Species"
  ) +
  theme_bw() +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",     # optional if still mapped
    "PILA" = "lightslateblue"
  )) +
  scale_x_continuous(
    breaks = seq(floor(min(trait_data1$elev_indiv)),
                 ceiling(max(trait_data1$elev_indiv)),
                 by = 500)
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

e3


```




# Figure 4 - HSM 
```{r}

h1 <- ggplot(data = subset(trait_data2, !type %in% c("HSM_predawn_2025", "HSM_midday_2025")),
       aes(x = diameter_complete, y = water_potential, color = type)) +
  facet_wrap(~ factor(species, levels = c("PILA", "PIPO", "ABCO", "ABMA")), nrow = 4) +
  geom_line(data = subset(trait_data2, type %in% c("midday_MPa_2025", "P50_mean")),
            aes(x = diameter_complete, y = water_potential, group = tree_ID),
            color = "gray40",
            size = 0.6,
            inherit.aes = FALSE) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c(
    "P50_MPa" = "blue",
    "P50_mean" = "darkblue",
    "P50_complete" = "darkgray",
    "midday_MPa_2025" = "darkcyan",
    "predawn_MPa_2025" = "steelblue3"
  )) +
  theme_bw() +
  labs(
    x = "Diameter at root collar (cm)",
    y = "Water Potential (-MPa)",
    color = "type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
h1


h2 <- ggplot(data = subset(trait_data2, !type %in% c("HSM_predawn_2025", "HSM_midday_2025")),
       aes(x = elev_indiv, y = water_potential, color = type)) +
  facet_wrap(~ factor(species, levels = c("PILA", "PIPO", "ABCO", "ABMA")), nrow = 4) +
  geom_line(data = subset(trait_data2, type %in% c("midday_MPa_2025", "P50_mean")),
            aes(x = elev_indiv, y = water_potential, group = tree_ID),
            color = "gray40",
            size = 0.6,
            inherit.aes = FALSE) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = seq(floor(min(trait_data1$elev_indiv)), ceiling(max(trait_data1$elev_indiv)), by = 500)) +
  scale_color_manual(values = c(
    "P50_MPa" = "blue", 
    "P50_mean" = "darkblue",
    "midday_MPa_2025" = "darkcyan",
    "predawn_MPa_2025" = "steelblue3"
  )) +
  theme_bw() +
  labs(
    x = "Elevation (ft)",
    y = "Water Potential (-MPa)",
    color = "type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

h2

h3 <- ggplot(data = subset(trait_data2, !type %in% c("HSM_predawn_2025", "HSM_midday_2025")),
       aes(x = percent_cover, y = water_potential, color = type)) +
  facet_wrap(~ factor(species, levels = c("PILA", "PIPO", "ABCO", "ABMA")), nrow = 4) +
  geom_line(data = subset(trait_data2, type %in% c("midday_MPa", "P50_mean")),
            aes(x = percent_cover, y = water_potential, group = tree_ID),
            color = "gray40",
            size = 0.6,
            inherit.aes = FALSE) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c(
    "P50_MPa" = "blue",
    "P50_mean" = "darkblue",
    "midday_MPa_2025" = "darkcyan",
    "predawn_MPa_2025" = "steelblue3"
  )) +
  theme_bw() +
  labs(
    x = "Canopy closure (%)",
    y = "Water Potential (-MPa)",
    color = "type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
h3

grid.arrange(h1,h3,h2, nrow = 1, ncol = 3)



```
# Figure 5 - growth 

```{r}

g1 <- ggplot(trait_data2, aes(x = diameter_complete, y = internode_length_1_2, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  labs(x = "Diameter at root collar (cm)", y = "Internode length (cm)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
  legend.position = "none"
  )

g1 

g2 <- ggplot(trait_data1, aes(x = predawn_MPa_combined, y = internode_length_1_2, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  labs(x = "Predawn water potential (-MPa)", y = "Internode length (cm)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "none"
  )

g2


g3 <- ggplot(trait_data2, aes(x =HSM_midday_P50_mean_2025 , y = trait_data2$internode_length_0_1, color = species)) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +  
  labs(
    x = "HSM midday (-MPa)",
    y = "Internode length (cm) ",
    color = "Species"
  ) +
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
  legend.position = "none"
  )

g3

g4 <- ggplot(trait_data2, aes(x = P50_complete, y = internode_length_1_2, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  labs(x = "P50 (-MPa)", y = "Internode length (cm)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "none"
  )

g4

g5 <- ggplot(trait_data2, aes(x = percent_cover, y = internode_length_1_2, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  labs(x = "Canopy closure (%)", y = "Internode length (cm)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "none"
  )

g5

g6 <- ggplot(trait_data2, aes(x = elev_indiv, y = internode_length_1_2, color = species)) +
  geom_point(size = 3, alpha = 1) +  
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +  
  scale_color_manual(values = c(
    "ABCO" = "mediumturquoise",
    "PIPO" = "lawngreen",
    "ABMA" = "maroon3",
    "PILA" = "lightslateblue"
  )) +
  labs(x = "Elevation (ft)", y = "Internode length (cm)") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "none"
  )

g6
grid.arrange(g1, g2, g3, g4, g5,g6, nrow = 3, ncol = 2)


```



```{r}
library(ggplot2)

ggplot(trait_data2, aes(x = DBH_complete, y = water_potential)) +
  geom_point(aes(shape = type, color = HSM_predawn_P50_mean), size = 3) +
  scale_shape_manual(values = c("predawn_MPa" = 16, "P50_mean" = 17, "P50_MPa" = 17)) +  # circle, triangle
  scale_color_gradient(low = "red", high = "blue", name = "HSM\n(P50 - predawn)") +
  facet_wrap(~species, ncol = 1) +
  theme_bw() +
  labs(
    x = "Diameter at root collar (cm)",
    y = "Water Potential (-MPa)",
    shape = "Type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )



library(dplyr)
library(tidyr)

# Create wide format with HSM
HSM_color_gradient <- trait_data2 %>%
  filter(type %in% c("predawn_MPa", "P50_mean")) %>%
  select(tree_ID, species, DBH_complete, type, water_potential) %>%
  pivot_wider(names_from = type, values_from = water_potential) %>%
  mutate(HSM = P50_mean - predawn_MPa)


library(ggplot2)

ggplot() +
  # Colored connecting lines
  geom_segment(data = HSM_color_gradient,
               aes(x = DBH_complete, xend = DBH_complete,
                   y = predawn_MPa, yend = P50_mean, color = HSM),
               size = 1) +
  
  # Points for both predawn and P50
  geom_point(data = HSM_color_gradient,
             aes(x = DBH_complete, y = predawn_MPa), 
             color = "steelblue", size = 3) +
  
  geom_point(data = HSM_color_gradient,
             aes(x = DBH_complete, y = P50_mean), 
             color = "darkblue", size = 3, shape = 17) +
  
  scale_color_gradient(low = "red", high = "blue", name = "HSM\n(P50 - predawn)") +
  facet_wrap(~species, ncol = 1) +
  theme_bw() +
  labs(
    x = "Diameter at root collar (cm)",
    y = "Water Potential (-MPa)"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


# Build polygons: x stays same, y goes from predawn to P50, loop back
HSM_polygons <- HSM_color_gradient %>%
  rowwise() %>%
  mutate(
    x1 = DBH_complete - 0.2,  # small width to create polygon
    x2 = DBH_complete + 0.2,
    y1 = predawn_MPa,
    y2 = P50_mean
  ) %>%
  ungroup() %>%
  rowwise() %>%
  do(data.frame(
    tree_ID = .$tree_ID,
    species = .$species,
    HSM = .$HSM,
    x = c(.$x1, .$x2, .$x2, .$x1),
    y = c(.$y1, .$y1, .$y2, .$y2)
  )) %>%
  ungroup()


ggplot() +
  geom_polygon(data = HSM_polygons, 
               aes(x = x, y = y, group = tree_ID, fill = HSM),
               alpha = 0.5, color = NA) +

  geom_point(data = HSM_color_gradient, 
             aes(x = DBH_complete, y = predawn_MPa), 
             color = "steelblue", size = 3) +

  geom_point(data = HSM_color_gradient, 
             aes(x = DBH_complete, y = P50_mean), 
             color = "darkblue", shape = 17, size = 3) +

  scale_fill_gradient(low = "red", high = "blue", name = "HSM\n(P50 - predawn)") +

  facet_wrap(~species, ncol = 1) +
  theme_bw() +
  labs(
    x = "Diameter at root collar (cm)",
    y = "Water Potential (-MPa)"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )



```


# stats - Table 1, 2, 3
```{r}

mod <- lmer(internode_length_1_2 ~ HSM_predawn_P50_mean_scaled + (1 | plot), data = ABMA)
summary(mod)



mod1 <- lmer(predawn_MPa ~ diameter + percent_cover + year + elev_indiv + species + (1|plot), data = trait_data2)
summary(mod1)

t.test(midday_MPa ~ year, data = trait_data2, paired = FALSE)
# There is a statistically significant difference between 2024 and 2025 predawn and midday water potential means. On average, predawn WP is higher (more negative) in 2024 and midday is 


mod2 <- lmer(HSM_predawn_P50_mean ~ DBH_complete_scaled + elev_indiv_scaled + percent_closure_scaled  + (1|plot), data = PIPO)
summary(mod2)

mod3 <- lmer(HSM_predawn_P50_mean ~ DBH_complete_scaled + percent_closure_scaled + elev_indiv_scaled + (1|plot), data = PIPO)
summary(mod3)


# scale predictor variables
trait_data1$diameter_complete_scaled <- as.numeric(scale(trait_data1$diameter_complete))
trait_data1$diameter_cm_scaled <- as.numeric(scale(trait_data1$diameter_cm))
trait_data1$percent_closure_scaled <- as.numeric(scale(trait_data1$percent_cover))
trait_data1$elev_indiv_scaled <- as.numeric(scale(trait_data1$elev_indiv))

# scale response variables 
trait_data1$predawn_scaled <- as.numeric(scale(trait_data1$predawn_MPa))
trait_data1$midday_scaled <- as.numeric(scale(trait_data1$midday_MPa))
trait_data1$P50_MPa_scaled <- as.numeric(scale(trait_data1$P50_MPa))
trait_data1$HSM_predawn_scaled <- as.numeric(scale(trait_data1$HSM_predawn))
trait_data1$HSM_predawn_P50_mean_scaled <- as.numeric(scale(trait_data1$HSM_predawn_P50_mean))
trait_data1$HSM_midday_P50_mean_scaled <- as.numeric(scale(trait_data1$HSM_midday_P50_mean))

# run models with scaled response and predictor variables
mod5 <- lmer(HSM_predawn_P50_mean ~DBH_complete_scaled * percent_closure_scaled * elev_indiv_scaled + (1|plot), data = PIPO)
summary(mod5)

install.packages("emmeans")
library(emmeans)

mod4 <- lmer(HSM_midday_P50_mean ~ DBH_complete_scaled * elev_indiv_scaled * percent_closure_scaled + (1|plot), data = PIPO)
summary(mod4)

# table 3 pairwise comparisons using emmeans
emms1 <- emmeans(mod4, pairwise ~ species)
summary(emms1$emmeans)
summary(emms1$contrasts) #automatically adjusts to tukey, for multiple comparisons



model <- lm(internode_length_0_1 ~ diameter_cm, data = trait_data1)
trait_data1$residuals <- resid(model)





```


# AICc model selection
```{r}
install.packages('MuMIn')
library(MuMIn)

dredge_df <- na.omit(ABCO[, c("HSM_midday", "elev_indiv", "DBH_complete", "percent_cover", "plot")])
global <- lmer(HSM_midday ~ elev_indiv * DBH_complete * percent_cover + (1|plot), data = dredge_df, na.action = na.fail)
dredge(global)

# likelihood ratio test for comparison of models by hypothesis testing versus AICc model selection
mod6<- lmer(HSM_predawn_P50_mean_scaled ~ 1 + (1|plot), data = dredge_df)
mod7 <- lmer(HSM_predawn_P50_mean_scaled ~elev_indiv_scaled + (1|plot), data = dredge_df)
mod8 <- lmer(HSM_predawn_P50_mean_scaled ~DBH_complete_scaled + elev_indiv_scaled + (1|plot), data = dredge_df)
mod9 <- lmer(HSM_predawn_P50_mean_scaled ~DBH_complete_scaled + elev_indiv_scaled + percent_closure_scaled +  (1|plot), data = dredge_df)

anova(mod6, mod7, mod8, mod9)
```

# fitted models
```{r}
ABCO_cleaned_data <- ABCO[!is.na(ABCO$DBH_complete) & !is.na(ABCO$percent_cover) & !is.na(ABCO$y), ]
ABCO_cleaned_data <- ABCO[!is.na(ABCO$DBH_complete) & !is.na(ABCO$percent_cover), ]


mod7<- lmer(ABCO$midday_MPa~ DBH_complete + percent_cover +DBH_complete:percent_cover + (1|plot), REML = TRUE, data = PILA)
summary(mod7)

#Get the fitted values (predicted values) from the model
fitted_values <- fitted(mod7)

# Create a new dataframe with predictors and fitted values
fitted_data <- data.frame(DBH = ABCO_cleaned_data$DBH_complete, 
                        percent_closure = ABCO_cleaned_data$percent_cover, 
                        fitted_values = fitted_values)

library(ggplot2)
ggplot(plot_data, aes(x = x1, y = fitted_values, color = x2)) +
  geom_line() +  # Fit line
  geom_point(data = your_data, aes(x = x1, y = y), color = "red", alpha = 0.5) +  # Raw data as points
  labs(title = "Fitted Values from Linear Mixed Model",
       x = "Predictor 1 (x1)", y = "Fitted Values (y)") +
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient for x2
  theme_minimal()

```


#climate check
```{r}
library(terra)
DATADIR = "/Users/anjumgujral/Box Sync/yuba-phys_data/climate-data"
# Site coordinates
site_coordinates <- data.frame(
  plot = paste("plot", c(7, 8, 9, 10, 11, 1, 2, 3, 4, 5)),
  lat = c(39.511469, 39.504699, 39.507502, 39.507644, 39.513851,
          39.491387, 39.490947, 39.487019, 39.478789, 39.472552),
  lon = c(-120.829953, -120.780825, -120.786236, -120.778930, -120.826322,
          -121.051720, -121.046958, -121.006063, -121.000819, -121.004718)
)

# Convert to spatial points
sites_sp <- vect(site_coordinates, geom = c("lon", "lat"), crs = "EPSG:4326")

# Load rasters (adjust paths to where you saved the files)

# Load the ASC file
aet_rast <- rast("/Users/anjumgujral/Documents/repos/yuba-phys/climate-data/BCMv8_1991_2020_aves/aet1991_2020_ave.asc")
pet_rast <- rast("/Users/anjumgujral/Documents/repos/yuba-phys/climate-data/BCMv8_1991_2020_aves/pet1991_2020_ave.asc")
map_rast <- rast("/Users/anjumgujral/Documents/repos/yuba-phys/climate-data/BCMv8_1991_2020_aves/ppt1991_2020_ave.asc")
cwd_rast <- rast("/Users/anjumgujral/Documents/repos/yuba-phys/climate-data/BCMv8_1991_2020_aves/cwd1991_2020_ave.asc")

# Extract PET and AET values for each plot
aet_vals <- extract(aet_rast, sites_sp)
pet_vals <- extract(pet_rast, sites_sp)
cwd_vals <- extract(cwd_rast,sites_sp)

# Combine into site data
site_coordinates$AET <- aet_vals[, 2]
site_coordinates$PET <- pet_vals[, 2]
site_coordinates$CWD <- cwd_vals[, 2]


ggplot(site_coordinates, aes(x = elevation, y = CWD, color = CWD)) +
  geom_point(size = 4) +
  scale_color_viridis_c() +
  labs(
    x = "Elevation (m)",
    y = "CWD",
    color = "CWD"
  ) +
  theme_minimal()

elevations <- c(
  5550,  # plot7
  5690,  # plot8
  5700,  # plot9
  5700,  # plot10
  5630,  # plot11
  3268,  # plot1
  3050,  # plot2
  4041,  # plot3
  4290,  # plot4
  4150   # plot5
)

```



# boxplots
```{r}
library(ggplot2)

ggplot(subset(trait_data2, !is.na(microsite) & !is.na(size)), aes(x = microsite, y = HSM_predawn_P50_mean, fill = size)) + 
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +
  geom_jitter(position = position_jitter(width = 0.2), color = "black", alpha = 0.6, size = 1.5) +
  #scale_fill_manual(values = c(
    #"seedling" = "skyblue",
    #"sapling" = "lightgreen"
  #)) +
  labs(
    x = "Microsite",
    y = "HSM (MPa)",
    fill = "Size"
  ) +
  facet_wrap(~ species) +  
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),  
    strip.text = element_text(size = 12)
  )

```
