---
title: "07_preliminary-analyses"
author: "Anjum K. Gujral"
date: '2024-11-11'
output:
  html_document:
    df_print: paged
  pdf_document:
    output_dir: /Users/anjumgujral/Documents/repos/yuba-phys
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
trait_data1 <- read.csv("/Users/anjumgujral/Documents/repos/yuba-phys/trait_data1.csv")

install.packages("tinytex")
library('tinytex')
library('dplyr')
library('tidyverse')
library('ggplot2')

```

Size distributions of seedlings and saplings

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = DBH_complete, fill = size_class)) +
  geom_histogram(binwidth = 0.5, position = "dodge", alpha = 0.7, color = "black") +
  labs( 
       x = "Diameter at root collar (cm)", 
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "lightgreen")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

Distribution of seedlings and saplings across light environments
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = DBH_complete, y = percent_cover, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
       y = "Canopy closure", 
       x = "Diameter at root collar") +
  scale_color_manual(values = c("blue", "green", "red", "purple")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

Distribution of microsite environments across elevation
-- how similar are microsites at high and low elevation 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = percent_cover, fill = microsite)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Canopy Closure"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))

# Create a new variable combining size_class and elevation
trait_data1$size_elevation <- interaction(trait_data1$size_class, trait_data1$elevation)


ggplot(trait_data1, aes(x = species, y = percent_cover)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Canopy closure", 
    x = "Species",
    color = "Size & Elevation"  
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )


ggplot(trait_data1, aes(x = microsite, y = percent_cover)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Canopy closure", 
    x = "Microsite level", 
    color = "Size & Elevation"  
  ) +
  facet_wrap(~ species, scales = "free_y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```

Distribution of stand basal area across elevation
-- how similar is stand density between plots at high and low elevation
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = stand_basal_area, fill = elevation)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Stand Basal Area"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))



ggplot(trait_data1, aes(x = species, y = stand_basal_area)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Stand Basal Area ", 
    x = "Species",
    color = "Size & Elevation"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )


ggplot(trait_data1, aes(x = microsite, y = stand_basal_area)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Stand Basal Area", 
    x = "Microsite level", 
    color = "Size & Elevation" 
  ) +
  facet_wrap(~ species, scales = "free_y") +
  
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```


```{r}

ggplot(data = trait_data1, aes(x = DBH_complete, y = midday_MPa, color = species)) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2, alpha = 0.7) +
  labs(
    x = "Diameter at root collar (cm)", 
    y = "Midday Water Potential (-MPa)", 
    color = "Species" 
  ) +
  
  theme_bw() +
  scale_color_manual(values = c("red", "blue", "green", "purple")) + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplot(data = trait_data1, aes(x = DBH_complete, y = midday_MPa, color = species, linetype = microsite)) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2, alpha = 0.7) +
  labs(
    x = "Diameter at root collar (cm)", 
    y = "Midday Water Potential (-MPa)", 
    color = "Species",  
    linetype = "Microsite"  
  ) +
  theme_bw() +
  scale_color_manual(values = c("ABCO" = "blue", "PIPO" = "green", "ABMA" = "red", "PILA" = "purple")) + 
  scale_linetype_manual(values = c("sun" = "dotted", "shade" = "solid", "int" = "twodash")) + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

```


Do seedlings and saplings differ in predawn and midday water potentials across elevation?

```{r}
ggplot(trait_data1, aes(x = DBH_complete, y = predawn_MPa, color = species)) +
  geom_point(size = 3) +                
  geom_smooth(method = "lm", se = FALSE, aes(group = species), linetype = "solid") +  
  labs(
    title = "Predawn Water Potential by Tree Diameter",
    x = "Diameter at root collar (cm)",
    y = "Predawn Water Potential (-MPa)",
    color = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "top")

```


```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = size_class, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())
```

```{r}
### stats
mod1 <- lm(predawn_MPa ~ species * size_class * elevation, 
             data = trait_data1)
summary(mod1)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = size_class, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod2 <- lm(midday_MPa ~ species * size_class * elevation, 
             data = trait_data1)
summary(mod2)

```

Do predawn and midday water potentials differ across microsite?
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod3 <- lm(predawn_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod3)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod4 <- lm(midday_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod4)

```

Do predawn and midday water potentials differ across elevation?
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Elevation",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod3 <- lm(predawn_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod3)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod4 <- lm(midday_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod4)

```

Do predawn and midday water potentials differ by microsite, species, and elevation?
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = microsite, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
       x = "Microsite",
       y = "Predawn Water Potential (-MPa)",
       fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
mod5 <- lm(predawn_MPa ~ species * microsite * elevation, 
             data = trait_data1)
summary(mod5)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
       x = "Microsite",
       y = "Midday Water Potential (-MPa)",
       fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
mod6 <- lm(midday_MPa ~ species * microsite * elevation, 
             data = trait_data1)
summary(mod6)
```

```{r}
library(lme4)
library(lmerTest)

ABCO <- trait_data1[trait_data1$species == 'ABCO',]
PILA <- trait_data1[trait_data1$species == 'PILA',]
PIPO <- trait_data1[trait_data1$species == 'PIPO',]
ABMA <- trait_data1[trait_data1$species == 'ABMA',]

mod7<- lmer(ABCO$midday_MPa~ microsite + size_class + microsite:size_class + (1|plot), REML = TRUE, data = ABCO)
summary(mod7)

```

# turning linear models into fitted models 
# run your lm() or gml() model
# The fitted values (predictions) are the values predicted by the model for each observation in the dataset based on the estimated model parameters. You can access them using the fitted() function or the $fitted.values attribute of the model object.





