---
title: "07_preliminary-analyses"
author: "Anjum K. Gujral"
date: '2024-11-11'
output:
  html_document:
    df_print: paged
  pdf_document:
    output_dir: /Users/anjumgujral/Documents/repos/yuba-phys
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
trait_data1 <- read.csv("/Users/anjumgujral/Documents/repos/yuba-phys/trait_data1.csv")

install.packages("tinytex")
library('tinytex')
library('dplyr')
library('tidyverse')
library('ggplot2')
install.packages("emmeans")
install.packages("lme4")
library(emmeans)
library(lme4)

```

Size distributions of seedlings and saplings

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = DBH_complete, fill = size)) +
  geom_histogram(binwidth = 0.5, position = "dodge", alpha = 0.7, color = "black") +
  labs( 
       x = "Diameter at root collar (cm)", 
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "lightgreen")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

# create a new column for size that categorizes individuals with diameter at root collar > 2.5 cm as saplings and individuals with diameter at root collar < 2.5 cm seedlings

trait_data1 <- trait_data1 %>%
  mutate(size = ifelse(DBH_complete < 2.5, "seedling", "sapling"))


```

Distribution of seedlings and saplings across light environments
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = DBH_complete, y = percent_cover, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
       y = "Canopy closure", 
       x = "Diameter at root collar") +
  scale_color_manual(values = c("blue", "green", "red", "purple")) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

Distribution of microsite environments across elevation
-- how similar are microsites at high and low elevation 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = percent_cover, fill = microsite)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Canopy Closure"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))

# Create a new variable combining size_class and elevation
trait_data1$size_elevation <- interaction(trait_data1$size, trait_data1$elevation)


ggplot(trait_data1, aes(x = species, y = percent_cover)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Canopy closure", 
    x = "Species",
    color = "Size & Elevation"  
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )


ggplot(trait_data1, aes(x = microsite, y = percent_cover)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Canopy closure", 
    x = "Microsite level", 
    color = "Size & Elevation"  
  ) +
  facet_wrap(~ species, scales = "free_y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```

Distribution of stand basal area across elevation
-- how similar is stand density between plots at high and low elevation
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = stand_basal_area, fill = elevation)) +
  geom_boxplot() +
  labs(
    x = "Elevation",
    y = "Stand Basal Area"
  ) +
    theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())+
  scale_fill_manual(values = c("skyblue", "lightgreen", "purple"))



ggplot(trait_data1, aes(x = species, y = stand_basal_area)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Stand Basal Area ", 
    x = "Species",
    color = "Size & Elevation"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )


ggplot(trait_data1, aes(x = microsite, y = stand_basal_area)) +
  geom_boxplot(alpha = 0.5, outlier.size = 3, outlier.colour = "black") +
  geom_point(aes(color = size_elevation), 
             position = position_jitter(width = 0.2, height = 0), 
             alpha = 0.7, size = 3) +
  labs(
    y = "Stand Basal Area", 
    x = "Microsite level", 
    color = "Size & Elevation" 
  ) +
  facet_wrap(~ species, scales = "free_y") +
  
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

```


```{r}
p2 <- ggplot(data = trait_data1, aes(x = DBH_complete, y = predawn_MPa, color = species, linetype = elevation)) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, alpha = 0.7) +
  labs(
    x = "Diameter at root collar (cm)", 
    y = "Predawn Water Potential (-MPa)", 
    color = "Species",  
    linetype = "Elevation"  
  ) +
  theme_bw() +
  scale_color_manual(values = c("ABCO" = "blue", "PIPO" = "green", "ABMA" = "red", "PILA" = "purple")) + 
  scale_linetype_manual(values = c("low" = "dotted", "high" = "solid")) + 
  scale_y_reverse() + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

mod1 <- lm(trait_data1$predawn_MPa~trait_data1$DBH_complete)
plot(trait_data1$DBH_complete, trait_data1$predawn_MPa, xlab="Diameter at root collar", ylab="Predawn water potential")
curve (coef(mod1)[1] + coef(mod1)[2]*x, add=TRUE) ## plotting the fitted regression line from the linear model rather than ggplot


trait_data2 <- trait_data1 %>%
  # Create predawn data
  mutate(time_of_day = "predawn", water_potential = predawn_MPa) %>%
  bind_rows(
    # Create midday data
    trait_data1 %>%
      mutate(time_of_day = "midday", water_potential = midday_MPa)
  )


p1 <- ggplot(data = trait_data2, aes(x = DBH_complete, y = water_potential, color = species, linetype = time_of_day)) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, alpha = 0.7) +
  labs(
    x = "Diameter at root collar (cm)", 
    y = "Water Potential (-MPa)", 
    color = "Species",  
    linetype = "Time of Day") +
  theme_bw() +
  scale_color_manual(values = c("ABCO" = "blue", "PIPO" = "green", "ABMA" = "red", "PILA" = "purple")) + 
  scale_linetype_manual(values = c("predawn" = "dotted", "midday" = "solid")) +
  scale_y_reverse() +  # Reverse y-axis for water potential
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

p3 <- ggplot(data = trait_data2, aes(x = DBH_complete, y = predawn_MPa, color = species, linetype = microsite)) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, alpha = 0.7) +
  labs(
    x = "Diameter at root collar (cm)", 
    y = "Predawn Water Potential (-MPa)", 
    color = "Species",  
    linetype = "Microsite") +
  theme_bw() +
  scale_color_manual(values = c("ABCO" = "blue", "PIPO" = "green", "ABMA" = "red", "PILA" = "purple")) + 
  scale_linetype_manual(values = c("sun" = "dotted", "int" = "twodash", "shade" = "solid")) +
  scale_y_reverse() +  # Reverse y-axis for water potential
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

p4 <- ggplot(data = trait_data2, aes(x = DBH_complete, y = midday_MPa, color = species, linetype = elevation)) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, alpha = 0.7) +
  labs(
    x = "Diameter at root collar (cm)", 
    y = "Midday Water Potential (-MPa)", 
    color = "Species",  
    linetype = "Elevation") +
  theme_bw() +
  scale_color_manual(values = c("ABCO" = "blue", "PIPO" = "green", "ABMA" = "red", "PILA" = "purple")) + 
  scale_linetype_manual(values = c("low" = "dotted", "high" = "solid")) + 
  scale_y_reverse() +  # Reverse y-axis for water potential
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
p5 <- ggplot(data = trait_data2, aes(x = DBH_complete, y = midday_MPa, color = species, linetype = microsite)) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, alpha = 0.7) +
  labs(
    x = "Diameter at root collar (cm)", 
    y = "Midday Water Potential (-MPa)", 
    color = "Species",  
    linetype = "Microsite") +
  theme_bw() +
  scale_color_manual(values = c("ABCO" = "blue", "PIPO" = "green", "ABMA" = "red", "PILA" = "purple")) + 
  scale_linetype_manual(values = c("sun" = "dotted", "int" = "twodash", "shade" = "solid")) +
  scale_y_reverse() +  # Reverse y-axis for water potential
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

install.packages("gridExtra")
library(gridExtra)
grid.arrange(p2, p4, p3, p4, p1, nrow = 3, ncol = 2)
```


Do seedlings and saplings differ in predawn and midday water potentials across elevation?

```{r}
ggplot(trait_data1, aes(x = DBH_complete, y = predawn_MPa, color = species)) +
  geom_point(size = 3) +                
  geom_smooth(method = "lm", se = FALSE, aes(group = species), linetype = "solid") +  
  labs(
    title = "Predawn Water Potential by Tree Diameter",
    x = "Diameter at root collar (cm)",
    y = "Predawn Water Potential (-MPa)",
    color = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "top")

```


```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = size, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())
```

```{r}
### stats
mod1 <- lm(predawn_MPa ~ species * size_class * elevation, 
             data = trait_data1)
summary(mod1)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = size_class, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
    x = "Size",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod2 <- lm(midday_MPa ~ species * size_class * elevation, 
             data = trait_data1)
summary(mod2)

```

Do predawn and midday water potentials differ across microsite?
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod3 <- lm(predawn_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod3)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod4 <- lm(midday_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod4)

```

Do predawn and midday water potentials differ across elevation?
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = elevation, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Elevation",
    y = "Predawn Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod3 <- lm(predawn_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod3)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  labs(
    x = "Microsite",
    y = "Midday Water Potential (-MPa)",
    fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
### stats
mod4 <- lm(midday_MPa ~ species * microsite, 
             data = trait_data1)
summary(mod4)

```

Do predawn and midday water potentials differ by microsite, species, and elevation?
```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = microsite, y = predawn_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
       x = "Microsite",
       y = "Predawn Water Potential (-MPa)",
       fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
mod5 <- lm(predawn_MPa ~ species * microsite * elevation, 
             data = trait_data1)
summary(mod5)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}

ggplot(trait_data1, aes(x = microsite, y = midday_MPa, fill = species)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +  
  facet_wrap(~ elevation) +
  labs(
       x = "Microsite",
       y = "Midday Water Potential (-MPa)",
       fill = "Species") +
  scale_y_reverse() +  # Reverse the y-axis
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())

```

```{r}
mod6 <- lm(elevation ~ species + microsite + elevation, 
             data = trait_data1)
summary(mod6)
```

```{r}
library(lme4)
library(lmerTest)

ABCO <- trait_data1[trait_data1$species == 'ABCO',]
PILA <- trait_data1[trait_data1$species == 'PILA',]
PIPO <- trait_data1[trait_data1$species == 'PIPO',]
ABMA <- trait_data1[trait_data1$species == 'ABMA',]

ABCO_cleaned_data <- ABCO[!is.na(ABCO$DBH_complete) & !is.na(ABCO$percent_cover) & !is.na(ABCO$y), ]
ABCO_cleaned_data <- ABCO[!is.na(ABCO$DBH_complete) & !is.na(ABCO$percent_cover), ]


mod7<- lmer(PIPO$midday_MPa~ percent_cover + DBH_complete + stand_basal_area + elevation +  DBH_complete:percent_cover + DBH_complete:elevation + (1|plot), REML = TRUE, data = PIPO)
summary(mod7)

# Get the fitted values (predicted values) from the model
fitted_values <- fitted(mod7)

# Create a new dataframe with predictors and fitted values
fitted_data <- data.frame(DBH = ABCO_cleaned_data$DBH_complete, 
                        percent_closure = ABCO_cleaned_data$percent_cover, 
                        fitted_values = fitted_values)

library(ggplot2)
ggplot(plot_data, aes(x = x1, y = fitted_values, color = x2)) +
  geom_line() +  # Fit line
  geom_point(data = your_data, aes(x = x1, y = y), color = "red", alpha = 0.5) +  # Raw data as points
  labs(title = "Fitted Values from Linear Mixed Model",
       x = "Predictor 1 (x1)", y = "Fitted Values (y)") +
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient for x2
  theme_minimal()

# Create a new dataframe that excludes species ABMA to look at elevation
trait_data2 <- trait_data1[trait_data1$species != "ABMA", ]

mod8<- lm(trait_data2$elev_band_ft ~ stand_basal_area + species + percent_cover + stand_basal_area:species + percent_cover:species, data = trait_data2)
summary(mod8) 

# Compute estimated marginal means for percent_cover
emm_percent_cover <- emmeans(mod7, ~ percent_cover)

# Compute estimated marginal means for DBH_complete (with percent_cover held constant)
emm_DBH <- emmeans(mod7, ~ DBH_complete)

# Alternatively, for interaction, you can compute both percent_cover and DBH_complete effects together
emm_interaction <- emmeans(mod7, ~ microsite * size)
summary(emm_interaction)
emm_result_df <- as.data.frame(emm_interaction)

# Plot the effect of percent_cover on midday_MPa
ggplot(emm_percent_cover, aes(x = percent_cover, y = emmean, ymin = lower.CL, ymax = upper.CL)) +
  geom_point() +
  geom_errorbar(width = 0.2) +  # Add error bars for confidence intervals
  labs(x = "Percent Cover", y = "Estimated Marginal Mean of Midday MPa") +
  theme_minimal() +
  ggtitle("Effect of Percent Cover on Midday MPa")

# Plot the interaction between percent_cover and DBH_complete
ggplot(emm_result_df, aes(x = size, y = emmean, color = microsite)) +
  geom_line() + 
  geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL, fill = microsite), alpha = 0.3) + 
  labs(x = "size", y = "Estimated Marginal Mean of Midday MPa") +
  #scale_color_viridis_c() +  # Color scale for percent_cover
  #scale_fill_viridis_c() +  # Color scale for the ribbon fill
  theme_minimal() +
  ggtitle("Interaction effect between microsite and size on midday WP")

# Inspect the structure of your emm_result_df
str(emm_result_df)

# Check the first few rows of the dataframe
head(emm_result_df)

# Summary statistics to see if values are within expected ranges
summary(emm_result_df)

```


# turning linear models into fitted models 
# run your lm() or gml() model
# The fitted values (predictions) are the values predicted by the model for each observation in the dataset based on the estimated model parameters. You can access them using the fitted() function or the $fitted.values attribute of the model object.





